Evolutionary unique N-glycan-dependent protein quality control system plays pivotal roles in cellular fitness and extracellular vesicle transport in Cryptococcus neoformans

Version 1

Abstract

Abstract

A conserved N-glycan-dependent endoplasmic reticulum protein quality control (ERQC) system has evolved in eukaryotes to ensure accuracy during glycoprotein folding. The human pathogen Cryptococcus neoformans possesses a unique N-glycosylation pathway that lacks the glucose addition step to the core N-glycan precursors in the ER but includes additional basidiomycetes-specific mannosidases. To investigate the molecular features and functions of the ERQC system in C. neoformans, we characterized a set of mutants with deletion of genes coding for the ERQC sensor UDP-glucose:glycoprotein glucosyltransferase (UGG1) and putative α1,2-mannose trimming enzymes (MNS1, MNS101, MNL1, and MNL2). The ugg1Δ, mns1Δ, mns101Δ, and mns1Δ101Δ mutants showed alterations in N-glycan profiles, defective cell surface organization, decreased survival in host cells, and varying degrees of reduced in vivo virulence. The ugg1Δ strain exhibited severely impaired extracellular secretion of capsular polysaccharides and virulence-related enzymes. Comparative transcriptome analysis showed the upregulation of protein folding, proteolysis, and cell wall remodeling genes, which is indicative of induced ER stress in ugg1Δ. However, no apparent changes were observed in the expression of genes involved in protein secretion or capsule biosynthesis. Additionally, extracellular vesicle (EV) analysis combined with proteomic analysis showed significant alterations in the number, size distribution, and cargo composition of EVs in ugg1Δ. These findings highlight the essential role of the functional ERQC system for cellular fitness under adverse conditions and proper EV-mediated transport of virulence bags, which are crucial for the full fungal pathogenicity of C. neoformans.

Introduction

Glycoproteins, destined for the secretory pathway, are synthesized on ribosomes attached to the surface of the endoplasmic reticulum (ER). They then enter the ER lumen where protein folding occurs (Rapoport, 2007). However, the accumulation of misfolded proteins affects cell viability and homeostasis; therefore, eukaryotes have evolved a conserved ER quality control (ERQC) system that recognizes folding defects, repairs them, and ensures the translocation of irreparable misfolded proteins into the cytosol for proteasome-mediated degradation via the ER-associated degradation (ERAD) system (Thibault and Ng, 2012; Xu and Ng, 2015; Balchin et al., 2016).

N-glycosylation is the most common post-translation modification, and the ERQC system relies on it to determine protein folding conformation (Aebi, 2013; Varki, 2017). Most eukaryotes synthesize a Dol-PP-linked Glc3Man9GlcNAc2 oligosaccharide as a common core N-glycan, which attaches to nascent polypeptides in the ER via an asparagine residue. Then, glucosidases I and II (Gls1 and Gls2) remove two glucose residues from the core oligosaccharide, after which, the glycoproteins are transported from the ER to the Golgi apparatus. Proteins containing the core monoglucosylated N-glycans (Glc1Man9GlcNAc2) enter a calnexin and/or calreticulin (CNX/CRT in mammals; CNE in yeast) chaperone-mediated folding cycle. Finally, Gls2 cleaves the remaining glucose residue. If the proteins are misfolded, they are recognized by the ERQC checkpoint enzyme, UDP-glucose:glycoprotein glucosyltransferase (UGGT), which reglucosylates them for re-entry into the folding cycle (Fig. 1A). Then, the N-glycans of the accurately folded proteins are further processed by Gls2 and α1,2-mannosidase I (Mns1) and moved to the Golgi apparatus. However, irreparably misfolded glycoproteins are targeted for ERAD. They undergo demannosylation and retro-translocation for proteasomal degradation in the cytosol. Recently, N-glycan precursors in the ER of some protists and fungi were found to be shorter than the typical 14-sugar N-glycan precursors in most eukaryote organisms. The length of these N-glycan precursors significantly impacts N-glycan-dependent QC of glycoprotein folding and ERAD (Banerjee et al., 2007; Samuelson et al., 2015).

Presence of UDP-glucose:glycoprotein glucosyltransferase (UGGT) and α1,2-mannosidases as endoplasmic reticulum protein quality control (ERQC) components in Cryptococcus neoformans

(A) Graphical representation of the ERQC pathway in Hommo sapiens, Saccharomyces cerevisiae, and Cryptococcus neoformans. (B) Domain structures of proteins encoded by C. neoformans UGG1 (CNAG_03648), MNS1 (CNAG_02081), MNS101 (CNAG_03240), MNL1 (CNAG_01987), and MNL2 (CNAG_04498). (C). Induced expression of ERQC genes in C. neoformans. Yeast cells were cultured in yeast extract peptone dextrose (YPD) medium to a mid-logarithmic phase and exposed to dithiothreitol (DTT; 20 mM). Tunicamycin (TM; 5 µg/ml) or to 37 °C for 1 h. The relative transcript levels of C. neoformans genes were analyzed using qRT-PCR. Transcript levels were normalized with that of ACT1. Error bars represent standard deviation of duplicated assays. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. *** P < 0.0005, ** P < 0.003, *** P < 0.005, * P < 0.05.

The basidiomycetous fungus Cryptococcus neoformans is an opportunistic, encapsulated human pathogen that primarily affects immunocompromised individuals, such as those with HIV/AIDS, causing fatal meningoencephalitis (Gottfredsson and Perfect, 2000; Kwon-Chung et al., 2000). The N-glycosylation pathway of C. neoformans is evolutionarily conserved; nevertheless, the structure and biosynthesis of its N-glycans shows several unique features (Park et al., 2012). C. neoformans contains serotype-specific high-mannose-type N-glycans with or without a β-1,2-xylose residue attached to the trimannosyl core. Additionally, the acidic N-glycans of C. neoformans contain xylose phosphates attached to the mannose residues both within the N-glycan core and outer mannose chains. The intact core N-glycan structure is crucial for C. neoformans pathogenicity (Thak et al., 2020); hence, alterations in the N-glycan structure modulates the interaction between the cell surface mannoproteins of C. neoformans and host cells (Lee et al., 2023). Unlike that observed in most eukaryotes, the N-glycosylation pathway in C. neoformans lacks homologous genes to ALG6, ALG8, and ALG10, which encode the glucosyltransferases that add three glucose residues to the core N-glycan before its attachment to proteins (Park et al., 2012). Man7GlcNAc2 and Man8GlcNAc2 without glucose residues are primarily detected in Dol-PP-linked glycans of C. neoformans (Samuelson et al., 2005). Moreover, the mature core N-glycan structures assembled on the cell surface mannoproteins of C. neoformans are primarily Man6–7GlcNAc2, which are shorter than the expected Man8GlcNac2 (Park et al., 2012). This observation led to the speculation that the terminal α-1,2-mannose residues of C. neoformans N-linked glycans may be more susceptible to trimming by ER-1,2 mannosidases due to the lack of glucose residues compared to those of most eukaryotes. Alternatively, the presence of multiple 1,2 mannosidases may generate more extensively trimmed core N-glycans in C. neoformans.

Cryptococcus neoformans uses several virulence factors to evade the host immune system and enhance its pathogenicity. These include an extensive polysaccharide capsule composed of glucuronoxylomannan (GXM) and galactoxylomannan (GXMGal) (Doering et al., 2009; Zaragozza et al., 2009), melanin (Qiu et al., 2012; Zhu and Williamson, 2004), and various extracellular enzymes such as phosphatase and urease (Singh et al., 2013). These essential virulence factors are primarily transported within the EVs in C. neoformans (Casadevall et al., 2019; Rizzo et al., 2021; Rodrigues et al., 2007). Therefore, C. neoformans EVs represent a heterogeneous population of “virulence bags” containing numerous fungal survival and pathogenicity-associated molecules. Additionally, recent proteomics analysis of EV cargos has identified several cell surface glycoproteins including members of the CDA family, which are well-known immunomodulators (Rizzo et al., 2021; Spetch et al., 2017).

In this study, we report for the first time the molecular features and function of N-glycan-dependent ERQC in C. neoformans through a systematic analysis of mutant strains lacking a set of ERQC gene homologs. Notably, our results revealed the pivotal roles of the ERQC system not only in maintaining cellular fitness but also in the EV-mediated transport of virulence factors in C. neoformans.

Results

Evolutionary unique features of C. neoformans ERQC components

We performed BLAST analysis of the C. neoformans H99 genome to identify the ERQC pathway-associated homologous genes of C. neoformans, followed by comparison with the ERQC components in other eukaryotic organisms (Supplementary Fig. S1, A). Unlike most eukaryotes, several yeast species within the Ascomycota phylum, such as Saccharomyces cerevisiae and Candida albicans, do not possess a functional UGGT. In contrast, the UGGT homologs were identified in most fungal species belonging to Basidiomycota, including C. neoformans (Supplementary Fig. S1, A and B). The C. neoformans UGGT (CNAG_03648), which has been named as Ugg1, consists of 1,582 amino acids (aa) and features a signal peptide (1–20 aa) along with four tandem-like thioredoxin-like (TRXL) domains: TRXL12 (33–320 aa), TRXL13 (28–416 aa), TRXL14 (432–618 aa), and TRXL15 (712–950 aa). All these domains show a highly conserved structural organization. Ugg1 also contains a glucosyltransferase (GT) 24 domain with a DXD motif (1,369–1,371 aa) and a KDEL-like ER retention signal (1579–1582 aa), which facilitates its retrieval from the Golgi apparatus (Fig. 1B, Supplementary Fig. S2, A).

Additionally, we identified two C. neoformans ORFs, Mns1 (CNAG_02081) and Mns101 (CNAG_03240), as homologs of the eukaryote α1,2-mannosidase I, which processes N-glycans before exporting them to the Golgi. The C. neoformans Mns1 and Mns101 show 42.4% and 28.8% amino acid identities to S. cerevisiae ER-α1,2-mannosidase I, respectively, and share 30.9% identity between them. Furthermore, C. neoformans Mnl1 (CNAG_01987) and C. neoformans Mnl2 (CNAG_04498) were identified as putative components of ERQC in C. neoformans (Supplementary Fig. S1, A and C). Whereas C. neoformans Mnl1 is a homolog of the yeast α1,2-mannosidase-like protein (Htm1), which processes N-glycans that are re-directed towards ERAD, C. neoformans Mnl2 encodes a mannosidase that does not show similarity to other eukaryotes mannosidases. Mns1, Mns101, and Mnl1 possess glucosyl hydrolase (GH) 47 domains, which are essential for mannosidase activity, whereas Mnl2 contains a GH92 domain, also known for its mannosidase activity (Fig. 1B). The Mns1 and Mnl1 protein families are characterized by the presence of a conserved cysteine (Cys) and alanine (Ala) residue (Jakob et al., 2001). These residues are located within the activity domain of cryptococcal Mns1, Mns101 and Mnl1, but not of Mnl2 (Supplementary Fig. S2, B). Notably, Mns101 and Mnl2 have been identified only in Basidiomycota yeasts (Supplementary Fig. S1, A) and seem to have diverged early from other mannosidase clades in the phylogenetic tree of fungal mannosidases (Supplementary Fig. S1, C). This indicates that Mns101 and Mnl2 are basidiomycete-specific proteins.

To examine the changes in ERQC-related gene expression under stress conditions, C. neoformans, which was cultivated in the presence of tunicamycin (TM, 5 µg/ml) or dithiothreitol (DTT, 20 mM) at 37 °C for 1 h, was subjected to quantitative reverse transcription-PCR (qRT-PCR) analysis (Fig. 1C). The analysis showed that KAR2, which encodes for a molecular chaperone associated with the unfolded protein response (UPR) system, was upregulated in response to inhibition of N-glycosylation by TM treatment, blockage of disulfide bond formation by DTT treatment, and heat stress at 37 °C. The ERQC components UGG1, MNS1, MNS101, MNL1, and MNL2 were upregulated by DTT treatment; however, they were not significantly induced after 1 h of TM treatment. Under high-temperature conditions, only UGG1 and MNL2 exhibited an induced expression pattern. Notably, DTT treatment, which results in rapid accumulation of misfolded proteins in the ER, induced a 10-fold higher expression of MNS1 and MNL1. These results strongly suggest that the C. neoformans genes UGG1, MNS1, MNS101, MNL1, and MNL2 are crucial components of the ERQC and ERAD systems.

Loss of UGG1, MNS1, and MNS101 causes alteration of N-glycan profiles in C. neoformans

We performed high-performance liquid chromatography (HPLC) analysis of the cell wall mannoproteins (cwMPs) from both the wild-type (WT) and ERQC mutant strains to investigate the ERQC malfunction-induced structural differences in N-glycans (Fig. 2). The HPLC profiles of cwMPs from the WT strain showed an M8 peak as the major species. This peak corresponded to N-glycans with 8 mannose residues (Fig. 2A, top). The glycan structure at the M8 peak (Man8GlcNAc2) in the WT primarily corresponded to the Man7GlcNAc2 core N-glycan with an additional mannose residue that is linked via an α1,6-linkage and added in the Golgi apparatus (Park et al., 2012). In the ugg1Δ mutant, M8 was also the main N-glycan species, but the pools of hypermannosylated N-glycans (larger than the M11 peak) were markedly reduced (Fig. 2A, middle). The altered N-glycan profile of the ugg1Δ mutant was restored to that of the WT after complementation with the WT UGG1 gene (Fig. 2A, bottom). Additionally, lectin blotting of the extracellularly secreted proteins from WT and ugg1Δ strains showed a distinct increase in the quantity of lower molecular weight secretory glycoproteins in the ugg1Δ strain compared with that of the WT (Fig. 2B). These results suggest that Ugg1 function is essential for the hypermannosylation of N-glycans that occurs in the Golgi apparatus.

High performance liquid chromatography (HPLC)-based N-glycan profiles of C. neoformans ERQC mutant strains

(A) Analysis of N-glycan profiles of the ugg1Δ mutant. (B) Silver staining and lectin blotting of sodium dodecyl sulphate (SDS)-polyacrylamide gels containing intracellular or secreted proteins into the culture supernatants of the wild type (WT) and ugg1Δ strains. Yeast cells were cultivated in YPD medium for 24 h, harvested, and subjected to sample preparation of soluble intracellular proteins and secreted proteins. The samples (30 μg, proteins) were loaded on 15% SDS-polyacrylamide gel and analyzed using silver staining (left) or blotting (right) with Galanthus nivalis agglutinin (Roche), which recognizes terminal α1,2-, α1,3-, and α1,6-linked mannose residues. (C) HPLC analysis of total N-glycan profiles of mns1Δ, mns101Δ, and mns1Δ101Δ mutants. (D) MALDI-TOF profiles of neutral N-glycans of mns1Δ, mns101Δ, and mns1Δ101Δ mutants. The N-glycans of cell wall mannoproteins from WT, ugg1Δ, ugg1Δ::UGG1, mns1Δ, mns101Δ, mns1Δ101Δ, mns1Δ::MNS1, mns1Δ::MNS101, mns1Δ101Δ::MNS1, and mns1Δ101Δ::MNS101 mutant strains were AA-labelled and analyzed using HPLC. Neutral N-glycan fractions were obtained from the HPLC fractionation of total N-glycans.

The mns1Δ mutant N-glycan profile showed a peak shift from an M8 to an M9 form, which strongly indicates that C. neoformans Mns1 could be the well-known ER α1,2-mannosidase I (Fig. 2C, mns1Δ). Notably, the loss of Mns101, which is present only in Basidiomycota, increased the fractions containing hypermannosylated glycans (> M10) while maintaining M8 as a primary core N-glycan form. This suggests that the basidiomycete-specific Mns101 may potentially be a novel α1,2-mannosidase that functions to remove mannose residues from hypermannosylated N-glycans in the Golgi apparatus or further trims the M8 glycan in the ER before the glycoproteins are transported to the Golgi (Fig. 2C, mns101Δ). The matrix-assisted laser desorption ionization-time of flight (MALDI-TOF) mass spectrometer analysis of neutral N-glycans, collected from HPLC fractionation (Fig. 2C), further confirmed the shift of the major M8 peak in the WT strain to M9 in the mns1Δ mutant and the increase in the hypermannosylation profile in the mns101Δ mutant (Fig. 2D). The N-glycan profiles of the mns1Δ101Δ double mutant showed the combined effect of each null mutation, and it showed both increased hypermannosylated glycans and a shift from the M8 to the M9 peak in both the HPLC (Fig. 2C, mns1Δ101Δ) and MALDI-TOF analyses. This indicates that Mns1 and Mns101 serve as mannosidases and play independent roles at different stages of N-glycan processing in C. neoformans.

In contrast, the loss of both MNL1 and MNL2 did not show notable differences in the HPLC profile of N-glycans from cell surface mannoproteins (Supplementary Fig. S3, A). Considering the expected function of Mnl1 and Mnl2 in ERAD, their substrates are mostly misfolded proteins that have not been transported to the Golgi apparatus. Thus, we speculate that only the normally processed N-glycan profiles of cell surface mannoproteins in the mnl1Δ mnl2Δ mutant strain were observed.

Loss of ERQC components results in defective growth fitness and increased stress sensitivity

We investigated the growth of the ugg1Δ mutant under various stress-inducing conditions to elucidate the roles of the ERQC components in ER stress response and adaptation, which are closely associated with C. neoformans virulence. The ugg1Δ mutant showed impaired growth even under normal growth conditions (Fig. 3A) and increased sensitivity to ER stress-inducing agents such as DTT, TM and cell wall stressors such as calcofluor white (CFW), Congo red (CR), sodium dodecyl sulfate (SDS), and caffeine. Additionally, ugg1Δ showed hindered growth in the presence of antifungal drugs such as azole agents (fluconazole and ketoconazole) and the glucose transport inhibitor fludioxonil. These findings suggest that Ugg1 is crucial for the robust growth and survival of C. neoformans under various stress conditions. In contrast, the single deletion strains (mns1Δ and mns101Δ) and double deletion strain (mns1Δ101Δ) did not exhibit noticeable phenotypic changes except for a slight increase in sensitivity of mns101Δ and mns1Δ101Δ to higher fludioxonil concentrations (Fig. 3B). Similarly, neither the single nor double disruption of MNL1 and MNL2 produced detectable changes in the tested growth phonotypes of C. neoformans cells (Supplementary Fig. S3, B).

Growth phenotype of C. neoformans ERQC mutant strains Spotting analysis of C. neoformans ugg1Δ

(A) and mns1Δ101Δ (B) mutant strains under various stress conditions such as heat stress (37 °C and 39 °C), ER stress (DTT: dithiothreitol, TM: tunicamycin), cell-wall stress (CFW: calcofluor white, CR: Congo red, SDS: Sodium dodecyl sulfate, caffeine), osmotic stress (NaCl, KCl, sorbitol) and treatment with antifungal drugs (fluconazole, ketoconazole, fludioxonil). (C) Growth analysis in the presence of 5’,5’,5’-trifluoroleucine (TFL). Respective strains were spotted on SC-Leucine media with or without TFL supplementation. Plates were incubated for 3 days at 30 °C. (D) RT-PCR analysis of IRE1-dependent splicing of HXL1. Strains were cultured in YPD supplemented with 5 µg/ml tunicamycin (TM). (E) Subcellular localization of GFP-tagged Ugg1, Mns1, and Mns101. ER tracker was used to visualize the ER membrane. Scale bar, 2.5 μm.

To determine whether the defective ugg1Δ phenotypes were induced partially by ER stress caused by the accumulation of misfolded proteins, we performed a growth analysis in the presence of 5’,5’,5’-trifluoroleucine (TFL), which is a leucine analog (Fig. 3C). We hypothesized that TFL incorporation into proteins could cause improper folding, which would ultimately negatively impact cell growth. ugg1Δ showed noticeably inhibited growth in the presence of TFL, whereas the WT strain showed no growth inhibition. The mns1Δ101Δ strain also exhibited slightly increased sensitivity to TFL compared with that of the WT strain. As a defense mechanism against misfolded protein accumulation-mediated ER stress, UPR is induced by the unconventional splicing of the HXL1 transcription factor in C. neoformans (Cheon et al., 2011). RT-PCR analysis showed that spliced HXL1 level was significantly higher in the ugg1Δ strain compared with that of WT, even under normal culture conditions (Fig. 3D). Moreover, enhanced HXL1 splicing in ugg1Δ strongly suggests that UGG1 loss results in misfolded protein accumulation, which causes ER stress even under normal growth conditions. Furthermore, green fluorescence protein (GFP)-tagged Ugg1, Mns1, and Mns101 proteins colocalized with the ER marker, confirming them to be functional ERQC components based on their subcellular localization in the ER (Fig. 3E).

ERQC defects lead to virulence attenuation in C. neoformans

We investigated the effects of a defective ERQC on in vitro virulence phenotypes of C. neoformans by analyzing cryptococcal capsule and melanin pigment production, as these are the two of the major virulence factors of C. neoformans. Culturing on L-DOPA-containing plates showed that melanin production was lower in the ugg1Δ strain compared with that in the WT (Fig. 4A). In contrast, the mns1Δ101Δ mutant cells did not show detectable defects in melanin production at either 30 or 37 °C (Fig. 4B). India ink staining of the capsule showed significantly reduced capsule thickness in the ugg1Δ strain and a moderate defect in the mns1Δ101Δ strain (Fig. 4C).

In vitro and in vivo virulence-associated phenotypes of C. neoformans UGG1, MNS1, and MNS101 mutant strains

(A, B) Melanin synthesis. WT, ugg1Δ, ugg1Δ::UGG1, mns1Δ, mns101Δ, mns1Δ101Δ, mns1Δ101Δ::MNS1, mns1Δ101Δ::MNS101, and cac1Δ (negative control) strains were serially diluted, plated on L-DOPA plates, and incubated at 30 °C and 37 °C. (C) Capsule formation. Cells were cultured for 2 days in 10% Sabouraud media at 30 °C and observed under the microscope. Statistical significance: ****, P < 0.0001, ns, not significant.). (D) In vivo virulence analysis. A/Jcr mice (n=8) were infected with 105 cells of WT, ugg1Δ, and ugg1Δ::UGG1, mns1Δ, mns101Δ, mns1Δ101Δ, mns1Δ101Δ::MNS1, and mns1Δ101Δ::MNS101 strains, and survival was monitored for 2 months. (E) Survival of C. neoformans in macrophages. Survival of C. neoformans cells within the J774A.1 macrophage-like cell line was determined by counting colony formation unit (CFU) obtained from lysed macrophages from two biologically independent experiment sets. **** P < 0.0001, *** P < 0.0005, * P < 0.05, ns, not significant. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test.

Defects in the ERQC system led to a more apparent decrease in pathogenicity in a murine model of systemic cryptococcosis. Notably, the ugg1Δ strain was completely avirulent in vivo (Fig 4D), and the infected animals did not display any signs of illness. Furthermore, MNS1 and MNS101 single disruptions did not cause a detectable decrease in C. neoformans pathogenicity; however, their double disruption resulted in decreased virulence. Histopathological analysis of infected mice lungs showed significantly reduced lung colonization by both ugg1Δ and mns1Δ101Δ cells (Supplementary Fig. S4, A). Additionally, the fungal burden of ugg1Δ-infected animals showed a notable decrease in organ colonization at 60 days post-infection (dpi) (Supplementary Fig. S4, B). Analysis of the fungal burden at 7 dpi showed significantly reduced organ colonization of the mns1Δ101Δ mutant compared with that of the WT (Supplementary Figure S4, C). Next, we co-cultured the mutant strains with the macrophage-like cell line J447A.1 to investigate the effects of ERQC defects on cryptococcal survival within macrophages (Fig. 4E). Consistent with the in vivo survival data, the number of surviving ugg1Δ and mns1Δ101Δ cells within macrophages was significantly lower compared with that of the WT strain. These data prove that a functional ERQC system is critical for maintaining full in vivo pathogenicity and ensuring robust survival within host immune cells.

Loss of Ugg1 function manifests defects in extracellular transport of GXM

Cryptococcal capsule materials are synthesized intracellularly and secreted to the cell surface. Subsequently, they are assembled and bound to the cell wall (Yoneda and Doering, 2006). The observed capsule defects in the ugg1Δ and mns1Δ101Δ mutant strains may be attributed to issues in the capsule polysaccharide synthesis or transport steps. Alternatively, defective capsule formation may result from increased shedding of capsule polymers without proper attachment to the cell surface. To determine the cause of the capsule defects, we examined the amounts of capsular polysaccharides synthesized intracellularly and those shed into the culture supernatant by a capsule blotting assay. We used the GXM-specific antibody 18B7 to analyze the WT and ERQC mutant strains, together with the cap59Δ and rim101Δ strains, which exhibit capsule formation defects. GXM is a major capsule component comprising α1,3-linked mannose residues with xylosyl and glucuronyl side groups. The cap59Δ strain lacks a visible capsule because of a defect in GXM synthesis (Grijpstra et al., 2009). C. neoformans rim101Δ exhibits a hypocapsular phenotype because of defective capsule attachment, despite normal polysaccharide synthesis and secretion (O’Meara et al., 2010). Intracellular GXM quantities did not notably differ between the WT, ugg1Δ, mns1Δ101Δ, and rim101Δ strains, although ugg1Δ showed slightly reduced GXM level (Fig. 5A, left). In contrast, no intracellular GXM was detected in cap59Δ, which was consistent with its defect in GXM synthesis. Moreover, the ugg1Δ strain did not show shedding of capsule polymers, whereas the mns1Δ101Δ strain showed shedding of GXM-containing polymers that were shorter in length than the WT-type polymers. This indicates poor polysaccharide polymerization (Fig. 5A, right). Reduced capsule thickness in mns1Δ101Δ is consistent with previous reports that capsule size is mediated at the polymer level (Yoneda et al., 2008). Therefore, these results suggest that ugg1Δ contains a defect in polysaccharide trafficking to the extracellular space, whereas mns1Δ101Δ contains a defect in polysaccharide polymerization, which results in defective capsule elaboration, albeit with differential degrees of severity in both ERQC mutants.

Capsule shedding and transfer analysis of C. neoformans UGG1, MNS1, and MNS101 mutant strains

(A) Capsule shedding analysis. Presence of intracellular (left) and shed (right) glucuronoxylomannan (GXM) was assessed by blotting a cell culture filtrate using the monoclonal antibody 18B7. The arrow indicates the direction of electrophoresis. (B) Capsule transfer analysis using exogenous capsule material from WT. The capsule transfer assay was performed using the indicated strains as acceptors. Surface capsules were probed using the anti-GXM antibody 18B7 conjugated with AlexaFluor 488. Quantitative measurement of fluorescence intensity was calculated based on two independent triplicate experiments with standard deviations presented as error bars. (C) Capsule transfer analysis using exogenous capsule material from WT or mns1Δ101Δ. Statistical significance: **** P < 0.0001. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. (D) Transmission electron microscopy (TEM) of C. neoformans WT, ugg1Δ, and mns1Δ101Δ strains. Yeast cells were grown overnight at 30 °C in YPD medium and fixed in 2% glutaraldehyde and 2% paraformaldehyde. A Zeiss Axioscope (A1) equipped with an AxioCan MRm digital camera was used to visualize India ink-stained C. neoformans cells. Specimens were prepared using critical point drying prior to TEM microscopy. Capsule and yeast cell body diameters were measured using ImageJ (National Institute of Health).

Furthermore, to determine whether the ERQC mutants are also impaired in capsule attachment, we performed a capsule transfer assay, which is a technique that assesses whether exogenously shed capsule polysaccharides bind to acapsular mutants (Reese and Doering, 2003). The hypocapsular rim101Δ mutant did not show reassociation of capsular polysaccharides; however, the acapsular cap59Δ mutant showed recovery of capsule formation when co-incubated with exogenous GXM (Fig. 5B). Similarly, ugg1Δ reverted its acapsular phenotype on incubation with the WT-shed polysaccharides, which strongly suggests that the defective capsule phenotype of ugg1Δ is primarily due to defective GXM synthesis or secretion rather than impaired attachment of capsule polymers. We further conducted a capsule transfer assay using capsular polysaccharides shed from mns1Δ101Δ. The cap59Δ mutant reverted to the capsular phenotype by attaching to the mns1Δ101Δ-shed polymers. However, the fluorescence intensity of the capsule generated by the mns1Δ101Δ-shed polymers was significantly lower than that of the cells attached with polysaccharides obtained from the WT. This result suggests that the mns1Δ101Δ double mutant secretes incomplete capsule polysaccharides (Fig. 5C), which leads to a hypocapsular phenotype. These results collectively indicate that the defective GXM trafficking to the extracellular space is caused by disruptive ERQC, impairing capsule formation.

Transmission electron microscopy (TEM) imaging of the WT and mutant cells distinctly showed a diminished capsule structure and loss of capsule shedding in the ugg1Δ strain compared with that of the WT strain (Fig. 5D). Additionally, we observed considerable thinning of the cell walls in both ERQC mutant cells. Notable changes in intracellular structures along with an increase in the number of pigmented vesicles were observed in the mutant cells even under YPD culture conditions. Under capsule-inducing conditions, a significant accumulation of vesicular structures (electron-lucent structures) was observed inside the cells, particularly in ugg1Δ. We performed LD staining to determine whether these vesicular structures might be lipid droplets (LDs). We observed a notable increase in LDs in ugg1Δ under capsule-inducing culture conditions (Supplementary Fig. S5, A). LDs impact proteostasis by sequestering misfolded proteins intended for degradation and providing an “escape hatch” when the ERQC is overloaded (Ploegh, 2007; Vevea et al., 2015). Additionally, FM4-64 dye staining showed increased number of vacuoles in the ugg1Δ cells compared with that in the WT (Supplementary Fig. S5, B). Moreover, all the observed abnormal phenotypes were less pronounced in the mns1Δ101Δ mutant than in the ugg1Δ strain. Taken together, the altered vesicular structures observed in the ugg1Δ and mns1Δ101Δ strains indicate that the ERQC function is crucial for maintaining proper vesicular structure and cell surface organization as both these functions are closely associated with the extracellular trafficking of virulence factors and cell wall remodeling.

ugg1Δ transcriptomic profiling shows induced ER and cell wall integrity stress responses

To elucidate the mechanisms underlying UGG1 deletion-induced physiological changes in C. neoformans, we performed RNA sequencing (RNAseq)-based transcriptome analysis of the WT and ugg1Δ strains under standard growth conditions (YPD medium, 30 °C). Differential gene expression analysis was performed with focus on changes exceeding 2-fold between the WT and ugg1Δ mutant strains (Fig. 6A). Comparative transcriptome analysis showed significant alterations in the expression patterns of 146 genes, of which 85 were upregulated and 61 were downregulated compared with that of the WT strain (Fig. 6B, Supplementary Tables S2, A and B).

Transcriptome analysis of C. neoformans WT and ugg1Δ cells

(A) Volcano plot comparing a 2-fold differential gene expression between ugg1Δ and WT strains under normal growth conditions. (B) Number of genes upregulated and downregulated by ≥ 2-fold in ugg1Δ compared with that of the WT. (C) Gene ontology (GO) analysis of differentially expressed genes between WT and ugg1Δ strains. Significantly upregulated genes in ugg1Δ are shown in red, whereas significantly downregulated genes in ugg1Δ are shown in blue. Total RNA was extracted and subjected to RNAseq analysis as described in the Supplementary Information. (D) qRT-PCR analysis of mRNA expression levels of a set of genes responsible for capsule biosynthesis, cell wall remodeling, and both conventional and non-conventional secretion in ugg1Δ vs WT under normal growth conditions from three biologically independent experiment sets.

Notably, the ugg1Δ mutant showed upregulation of genes such as SKN1 and KRE6 (transmembrane glucosidases involved in the sphingolipid biosynthesis and β-glucan biosynthesis), CAT2 (a putative peroxisomal catalase), CHS7 (chitin synthase export chaperone), and CNAG_05458 (a putative endo-1,3(4)-β-glucanase). In contrast, genes encoding ERG3 and ERG25 (associated with the ergosterol biosynthesis pathway) and FHB1 (flavohemoglobin denitrosylase associated with counteracting nitric oxide stress) were significantly downregulated. Gene Ontology (GO) analysis in ugg1Δ showered pronounced induction of genes implicated in various cellular processes, including the hydrolysis of O-glycosyl compounds for cell wall remodeling, DNA replication, proteolysis, ribosome biogenesis, protein folding, and serine/threonine kinase activity. In contrast, the genes associated with carbohydrate and lipid metabolism, chaperone binding, iron homeostasis, and mitochondrial intermembrane space were considerably downregulated (Fig. 6C). The transcriptome profile strongly suggests that loss of Ugg1 function induces the expression of several genes involved in maintaining cell wall integrity to compensate for cell wall defects, particularly those associated with chitin (CDA2, CHS7, QRI1, and CNAG_06898) and glucan biosynthesis (SKN1, KRE6, EBG1, LPI9, CNAG_05458, and BLG2). Furthermore, genes coding for ER chaperones that aid in protein folding (ERO1, KAR2, LHS1, and PDI1) and chaperone regulator (SCJ1), and genes involved in proteolysis in the ER (CNAG_04635, CNAG_06658) were induced, likely as part of the ER stress response induced by the accumulation of misfolded proteins in the presence of a defective ERQC.

Further investigation of the effects of UGG1 disruption in the expression of genes involved in capsule biosynthesis, cell wall remodeling, and the secretion pathways of protein virulence factors using qRT-PCR analysis (Fig. 6D) showed no noticeable repression of the expression of genes related to capsule biosynthesis and conventional or unconventional secretion pathways at the mRNA level, despite the aberrantly displayed defective phenotypes. Notably, ugg1Δ mutant showed notable upregulation of the genes involved in cell wall remodeling, as indicated by the GO analysis of transcriptome data. Taken together, these data strongly suggest that ERQC system malfunction caused by UGG1 disruption may exert negative effects on the production of virulence factors at post-transcriptional levels, most likely during the extracellular transport process.

EV-mediated protein trafficking and non-conventional secretion is impaired in the ugg1Δ mutant strain

Many enzymes contribute to the composite cryptococcal virulence phenotype. Some of these virulence-associated enzymes are secreted through traditional secretion pathways, whereas others are packaged into EVs and released into the extracellular milieu via non-conventional secretion mechanisms (Amelida et al., 2015). The defects observed in the ERQC mutant strains, particularly in the production of melanin and polysaccharide capsules, both of which are primarily transported by EVs, suggest a possible impairment in the EV-mediated extracellular transport of virulence-associated enzymes such as acid phosphatase, urease, and laccase (Rodrigues et al., 2008) in these mutants. Laccase enables melanin synthesis (Salas et al., 1996), urease hydrolyzes urea into ammonia, which facilitates C. neoformans invasion into the CNS (Olszewski et al., 2014), and acid phosphatase serves as an adhesion-related enzyme (Collopy-Junior et al., 2006). Therefore, we examined the in vitro activity of these virulence-related enzymes in both intracellular and extracellular secreted fractions.

Comparative analysis showed that whereas the changes in the intracellular activities of acid phosphatase, urease, and laccase were negligible, their activities were significantly low in the extracellular fractions of the ugg1Δ cells (Fig. 7A, 7B, 7C, and 7D). Urease analysis was performed in both solid and liquid media, and it showed a notable loss of secreted urease activity in ugg1Δ cells with a slight decrease in the mns1Δ101Δ double mutant cells (Fig. 7A and 7B). As urease lacks a signal peptide required for conventional protein secretion, low urease secretion strongly indicates a compromised EV-mediated non-conventional secretory pathway in the ugg1Δ mutant strain.

Analysis of protein secretion in C. neoformans UGG1, MNS1, and MNS101 mutant strains

(A) Spot assay for urease analysis on Christensen’s urea agar. Absence of pink coloration indicates loss of urease activity. (B, C, D) Analysis of secretion for virulence-related enzymes such as urease, laccase, and acid phosphatase from three biologically independent experiment sets. Statistical significance: **** P < 0.0001, ** P < 0.003, *** P < 0.005, * P < 0.05. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. (E, F) Analysis of secretion for non-virulence-related enzymes such as cellulase and α-amylase. Statistical significance: *** P<0.0005, ** P<0.003 * P < 0.05, ns, not significant. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. The enzymatic activities of urease, laccase, acid phosphatase, cellulase, and α-amylase in soluble cell lysates and culture supernatants were measured as described in the Materials and Methods section. (G) Analysis of the conventional secretion of Cda1 in C. neoformans. Presence of Cda1 was analyzed in total (T), soluble (S), and insoluble (I) fractions of intracellular extracts (left), along with the secreted fraction (right). Subcellular fractionations were performed as previously described (Thak et al., 2022), and the fractions were subjected to western blotting analysis using an anti-Cda1 antibody. (H) Analysis of raffinose utilization as the carbon source. Spotting analysis was performed on C. neoformans WT, ugg1Δ, and cap59Δ strains cultivated in SC media with or without glucose or raffinose supplementation. Plates were incubated for 3 days at 30 °C.

Next, we tested cellulase and α-amylase activities to assess the impact of disrupted ERQC system on the conventional secretion pathway of non-virulence-related enzymes (Fig. 7E and 7F). The activities of these carbohydrate polymer-degrading enzymes in the extracellular fraction of ugg1Δ were slightly lower than those of the WT; this likely indicates decreased intracellular production rather than significantly compromised secretion efficiency. Furthermore, we investigated the localization of chitin deacetylase I (Cda1), which is a glycosylphosphatidylinositol (GPI)-anchored protein that is involved in the conversion of chitin to chitosan for proper maintenance of cell wall integrity (Baker et al., 2007; Baker et al., 2011; Upadhya et al., 2021). Cda1 was predominantly detected in the insoluble cellular protein fraction, which includes the cell wall and membranes (Fig. 7G, left). This result is consistent with the localization of Cda1 on the cell surface via a GPI anchor. Cda1 is secreted extracellularly on cleavage of the GPI anchor. Notably, the ugg1Δ mutant strain shows much higher intracellular and extracellular Cda1 levels compared with that of the WT and mns1Δ101Δ strains (Fig. 7G, right). This concurs with the approximate 1.87-fold increase in CDA1 mRNA levels observed in the RNAseq data. Thus, the efficient surface localization and secretion of Cda1 strongly support the conclusion that the conventional secretion pathway remains functional without significant compromise in ugg1Δ.

The acapsular mutant cap59Δ strain was previously reported to have secretion defects, based on observations that the cap59Δ strain generated in the serotype D background (C. deneoformans) displayed defective secretion of enzymes involved in the hydrolysis of raffinose or urease (Garcia-Rivera et al., 2004). However, the cap59Δ strain used in this study, which was constructed from the serotype A strain H99 background (Thak et al., 2020), showed defective alkaline phosphatase secretion but only marginally defective urease secretion. The cap59Δ acapsular mutant constructed from the C. deneoformans background does not use raffinose as a carbon source because of poor protein secretion of raffinose hydrolyzing enzymes (Garcia-Rivera et al., 2004). Therefore, to test whether the ERQC-related mutant strains utilize raffinose as a carbon source, we performed a growth test using SC media containing raffinose as the sole carbon source. Notably, ugg1Δ and cap59Δ strains showed no evident defects in raffinose utilization compared with that of the WT (Fig. 7H). This suggests that the conventional secretion pathway is functional in both mutant strains. Overall, the results suggest that the UGGT-mediated ERQC defect exerts a more pronounced negative impact on EV-mediated protein transport and only a marginal effect on the conventional secretion pathway.

Extracellular vesicle biogenesis and cargo loading are defective in ugg1Δ strain

We analyzed the number and size distribution of EVs to determine if defects in the assembly and transport of EVs in the ugg1Δ mutant might cause abnormal EV-mediated trafficking of virulence factors. Nanoparticle tracking analysis (NTA) showed a major peak in size distribution at approximately 150 (134±28) nm in the WT strain, which was similar in size to mammalian exosomes and typical microbial EVs. Additionally, a minor peak ranging from 300–500 nm was observed and corresponded to the microvesicles (Fig. 8A). Notably, the size distribution of ugg1Δ EVs was more heterogeneous with smaller EVs ranging from 50–150 nm (82±16; 124±14) compared with that of the WT. The cap59Δ EVs displayed a major distribution of approximately 150 nm (120±21); however, the microvesicles were barely detected in either cap59Δ or ugg1Δ strains.

Analysis of extracellular vesicles purified from WT, ugg1Δ, and cap59Δ cells

(A, B) Nanoparticle tracking analysis (NTA) of EVs extracted from WT, ugg1Δ, and cap59Δ strains and quantification of total extracellular vesicle (EV) concentration per cell density. Quantitative measurements were derived from three independent experiments with standard deviations presented as error bars. Statistical significance: *** P<0.0005, * P< 0.05. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. (C, D) Cryo-TEM imaging of purified EVs and comparative analysis of EV size in WT, ugg1Δ, and cap59Δ strains. Scale bar, 100 nm. Total number of 100 EVs per strain, as captured using cryo-TEM, were analyzed for measuring the outer EV diameter. (E) Heatmap representation of fold change between WT and ugg1Δ EV-associated proteins, commonly detected in this study and in previously reported EV proteome datasets (ugg1Δ/WT). Upregulated proteins in ugg1Δ are shown in red, whereas downregulated proteins are shown in blue. (F) Blotting analysis of GXM in the C. neoformans cells and EVs of WT, ugg1Δ, and cap59Δ strains. The 8 M urea extracts were obtained from EVs and cell pellets, from which EVs are generated. The urea extracts (5 μg total proteins) were loaded on 8% SDS-polyacrylamide gel and subjected to silver staining or blotting analysis using the anti-GXM 18B7 (α-GXM) and anti-Cda1 (α-Cda1) antibodies, respectively. Left: total cell extract. Right: total EV extract.

We observed a significant reduction in the total number of secreted EVs in the ugg1Δ mutant (approximately 40%), which suggests defective EV biogenesis or stability in the absence of functional Ugg1 (Fig. 8B). Cryo-TEM analysis of EV morphology confirmed that the sizes of EVs released by ugg1Δ were significantly smaller and more diverse than those of the WT or the acapsular mutant (Fig. 8C); this further supports the NTA results. Examining EV size distribution by measuring their diameter (Fig. 8D) further confirmed that although the EVs in the WT strain ranged from approximately 50–550 nm in size with distribution primarily concentrated at approximately 150 nm, the EVs in the ugg1Δ mutant were smaller (< 100 nm) with heterogeneous sizes. Notably, cap59Δ EVs were uniformly distributed at approximately 150 nm and were present at a higher concentration than those of EVs from the WT. An increase in EV release and virulence factors has been also observed in acapsular mutant strains of C. neoformans and C. gattii (Rodrigues et al., 20007; Reis et al., 2019), which supports the notion that the capsule serves as a barrier in EV release in capsule forming species.

To examine the protein content in EVs from WT and ugg1Δ, EV-associated proteins were extracted in triplicate using 8 M urea from purified EVs, and the equivalent concentration of EV proteins were subjected to proteomics analysis (Fig. 8E; Supplementary Fig. S6, A). We identified 1,688 proteins and quantified 1,625 proteins (Supplementary Table S3). Among them, 1,008 proteins showed differences between WT and ugg1Δ (Supplementary Fig. S6, B). Compared with that of the WT EVs, the ugg1Δ EVs exhibited altered protein abundance with only 64 proteins showing upregulation and 777 proteins showing > 2-fold downregulation (Supplementary Fig. S6, C, D). Three proteomic analyses of C. neoformans EVs have been reported previously (Rodrigues et al., 2008; Wolf et al., 2014; Rizzo et al., 2021), identifying 76, 202, and 1,847 proteins associated with EVs in C. neoformans, respectively. Using the same culture medium and EV preparation method, the number and content of identified EV-associated proteins in our study was mostly similar to that reported by Rizzo et al., who further defined 71 non-ribosomal proteins as EV-enriched proteins after proteomic data enrichment analysis (Rizzo et al., 2021). We compared the protein content between WT and ugg1Δ among the EV proteins identified in our study with focus on the proteins commonly detected among either the 71 non-ribosomal EV-enriched proteins (Rizzo et al., 2021) or 50 overlapped proteins from two C. neoformans EV proteomic data sets reported by Rodrigues et al. and Wolf et al. (Bleackley et al., 2021). Among 59 EV-associated proteins that were commonly detected by previous and current studies, 24 proteins were detected at lower concentration, whereas 5 were present in higher concentrations of 2-fold difference in ugg1Δ mutant EVs compared with that of WT EVs (Fig. 8E; Supplementary Table S4). Thus, the proteome analysis showed that the absence of functional Ugg1 not only caused morphological alterations and reduced EV numbers but also altered EV protein abundance in C. neoformans.

To examine the loading defect of GXM in the ugg1Δ EVs, we examined the presence of GXM in purified EVs samples and compared the detected GXM quantities in the total cell extracts from WT, ugg1Δ, and cap59Δ strains (Fig. 8F). Cda1 was used as the representative EV-associated protein because the presence of Cda1 was reported within the C. neoformans EV membrane despite a substantial portion of Cda1 being secreted through the conventional secretion pathway (Rizzo et al., 2021). The 8 M urea extracts were obtained from the purified EVs and the whole cells that were subjected to pelleting during EV purification. The extracted samples, which contained equivalent protein concentrations, were subjected to blotting analysis using GXM-specific (18B7) and Cda1-specific antibodies (Fig. 8F). No GXM polysaccharides were detected in the EVs released from the ugg1Δ, although the whole-cell extract of ugg1Δ mutant strain contained a significant quantity of GXM. Taken together with the reduced quantity of EV-associated proteins, the absence of GXM polysaccharides in the ugg1Δ EVs strongly indicates that EV cargo loading is defective when the ERQC is dysfunctional in C. neoformans.

Discussion

As glycoproteins pass through the secretory pathway, which starts at the ER, their N-linked glycans undergo extensive modifications in association with protein quality control. In mammals, the ERQC system is composed of ER glucosyltransferases (ALG6, ALG8, and ALG10), ER glucosidases (GLS1 and GLS2), ER chaperones (calnexin and calreticulin, CNX/CRT), and the folding sensor UGGT. The addition and removal of glucose residues enables the substrates to undergo cycles of binding and release, which promote accurate folding and prevent incompletely folded proteins from exiting the ER. After glucose trimming, α1,2-mannosidase I (ERManI) in the ER cleaves a mannose residue at the B branch of the M9 oligosaccharides and converts the core M9 N-glycans (Man9GlcNAc2) to M8 forms (Man8GlcNAc2). This step occurs before the accurately folded proteins are exported to the Golgi apparatus or misfolded proteins are redirected towards degradation (Fig. 1A, Ziegler et al., 1991, Herscovics, 2001). Thus, in most eukaryotes, Man8GlcNAc2 is the final mature core N-glycan attached to glycoproteins in the ER before their transport to the Golgi complex for further N-glycan modification.

In contrast to higher eukaryotes, the ERQC system of S. cerevisiae is composed of glucosidase I, glucosidase II, and the calnexin Cne1 without a functional UGGT (Fig 1A; Supplementary Fig. 1A). Although Kre5 is homologous to UGGT, S. cerevisiae Kre5 functions are unrelated to the putative calnexin cycle in the ERQC, and the function of ER glucosidases I/II is primarily associated with the synthesis of cell wall β1,6-glucan (Fernandez et al., 1996; Simons et al., 1998). The ERQC composition of C. neoformans is unique in that it possesses UGGT but lacks the three glucosyltransferases necessary for adding glucose residues to the core precursor N-glycans (Park et al., 2012). This step is essential for generating the monoglucosylated form, which is recognized by CNE/CRT for initiating the folding mechanisms. Additionally, C. neoformans lacks CRT, which is the chaperone that aids CNE, and carries multiple α1,2-mannosidases. In this study, we investigated the molecular features and functions of the N-glycan-dependent ERQC in the human pathogen C. neoformans, which generates unique N-glycan precursors without glucose addition and shorter in length than those of most eukaryotes in the ER. Our data strongly suggest that despite the incomplete composition of the ERQC components, the UGGT-centered ERQC plays pivotal roles in cellular fitness, particularly in EV-mediated extracellular transport, which is crucial for pathogenicity.

The C. neoformans ugg1Δ mutant was avirulent in mice, a phenotype consistent with the observed defects in key virulence determinants, such as the polysaccharide capsule and melanin, as well as poor growth at 37 °C. The mns1Δ101Δ mutant displayed intermediate phenotypes between those of the WT and ugg1Δ. Our data from the UPR induction analysis and comparative transcriptomic analysis strongly indicate that the ERQC mutation generates ER stress; this accounts for significant similarity of the defective phenotypes of the C. neoformans ERQC mutants to those reported in other mutants that exhibit defective protein folding in the ER, particularly in terms of increased stress sensitivity and decreased virulence. Connections between ER stress and thermotolerance have previously been established in C. neoformans, as growth at 37 °C requires key ER protein chaperones and protein processing machinery, including components of the UPR signaling pathway Ire1 and the ER stress-responsive transcription factor Hxl1 (Cheon et al., 2011; Havel et al., 2011; Jung et al., 2013). Other mutations causing defects in ER function in C. neoformans, such as mutants lacking DNJ1 (an ER J-domain containing co-chaperone) and CNE1 (an ER chaperone), resulted in growth inhibition. The dnj1Δ mutant strain displayed impaired elaboration of virulence factors, such as the exopolysaccharide capsule and extracellular urease activity, when cultured at human body temperature (Horianopoulos et al., 2021). Taken together with our data from the ERQC mutants, these findings strongly support the notion that maintaining ER homeostasis is crucial for survival and virulence factor production at elevated temperatures.

It is quite notable that the C. neoformans ugg1Δ EVs exhibited defective loading of GXM and many protein cargos, alongside a decrease in the number and changes in the size distribution of the extracellular EVs. Combined with compromised cell fitness, these defects in EV biogenesis and cargo loading, even under normal growth conditions, likely contribute to the complete loss of virulence owing to the defective export of virulence factors in the ugg1Δ mutant. EV generation involves complex molecular networks that function cooperatively or independently to mediate the invagination and fission of membranes and selective sorting of cargo molecules. Fungal EVs are produced and released under the maturation of endosomes into multivesicular bodies (MVBs). These MVBs are targeted to the cell surface, where they fuse with the plasma membrane and release luminal MVB vesicles into the outer space, serving as an unconventional secretion pathway (Oliveira et al., 2010). MVB formation requires the functionality of the endosomal sorting complex required for transport (ESCR). The ESCRT pathway is highly complex and involves a series of finely regulated events (Henne et al., 2011). Deletion of several EV transport-related genes through the ESCRT complex directly impacts the cryptococcal capsule, primarily by reducing capsule size. Specifically, the C. neoformans vps27Δ strain, which is involved in the ESCRT complex, showed altered EV size distribution, reduced capsule dimensions, defects in laccase export to the cell wall, and poor extracellular export of urease (Park et al., 2020). Other regulators of unconventional secretion are also linked to EV biogenesis in C. neoformans. For example, the Golgi reassembly and stacking proteins (GRASPs) regulate EV cargo and dimensions (Peres et al., 2018). In C. neoformans, a graspΔ mutant strain produced EVs with dimensions that significantly differed from those produced by WT cells, along with attenuated virulence and abnormal RNA composition. Moreover, the GRASP protein Grasp homology 1 (GRH1) serves as a chaperone that directly influences EV cargo (Malhotra, 2013; Peres da Silva et al., 2018). Autophagy regulators, which participate in EV formation in other eukaryotes, also play a role in cryptococcal EV formation. An atg7Δ strain manifests hypovirulence, and EVs produced by this strain show slightly different RNA composition compared with that of the WT cells (Oliveira et al., 2016).

It is intriguing how the absence of UGGT, leading to ERQC defects, results in defective EV biogenesis and cargo loading in C. neoformans. The ER is one of the largest organelles in the cell and is involved in multiple fundamental biological processes such as protein folding and secretion, Ca2+ storage, and lipid synthesis. In addition to protein folding, the ER is crucial for regulating lipid metabolism (Moncan et al., 2020). Several enzymes involved in triglyceride (TG) biosynthesis, cholesterol biosynthesis, and enzymes regulating membrane turnover and dynamics are located in the ER. Consequently, ER stress significantly impacts lipid and sterol synthesis, although some of these mechanisms are yet to be clarified. LDs aid the UPR and ERAD in degrading misfolded proteins during ER stress in S. cerevisiae (Garcia et al., 2020). In the present study, we observed a drastic increase in LDs in the ugg1Δ mutant, consistent with the previous hypothesis that LDs help to maintain ER homeostasis. Furthermore, we observed increased number of vacuoles in ugg1Δ. The altered vesicular structures observed in the ugg1Δ and mns1Δ101Δ strains may indicate abnormal lipid homeostasis caused by the ERQC defects, which could, in turn, affect EV biogenesis. The importance of lipids and membrane regulators in proper EV formation and GXM export has been suggested in a previous study on the Apt1 flippase in C. neoformans (Rizzo et al., 2019). Additionally, ergosterol is essential for membrane fluidity, permeability, and protein transport (Ermakova and Zuev, 2017). Erg6 is an enzyme involved in the ergosterol biosynthesis pathway and was identified as essential for the trans-Golgi network transport of proteins (Proszynski et al., 2005; Nes et al., 2009). In C. neoformans, the erg6Δ mutant released EVs with a significantly larger diameter than those of the WT, carrying increased levels of proteins and sterols. This highlights the role of ergosterol in cryptococcal EV biogenesis (Oliveira et al., 2020). Taken together, these findings suggest that ER stress caused by misfolded glycoprotein accumulation in ERQC-defective mutants may alter lipid composition, which could affect EV biogenesis.

A recent study presented strong evidence for a key role of ER stress in modulating EV biogenesis by demonstrating that ER stress decreases exosome production through both adiponectin/T-cadherin-dependent and -independent pathways in mice (Fukuoka et al., 2023). They showed that T-cadherin is downregulated by ER stress through IRE1α activation at the mRNA and protein levels and that ER stress decreases EV production through adiponectin/T-cadherin-independent way, which might involve interferon pathway activation in mice. In this study, we similarly observed induced activity of Ire1 in C. neoformans ugg1Δ strain, even under normal culture conditions, which suggests a possible association between the Ire1-mediated UPR pathway and cryptococcal EV biogenesis (Fig. 9). Additionally, glycosylation regulates the biogenesis of small EVs and affects protein cargo loading efficiency (Harada et al., 2020). This suggests that the altered N-glycosylation observed in C. neoformans ugg1Δ may influence cargo loading of certain EV-targeting glycoproteins. Future studies are needed to investigate the ERQC-mediated modulation of EV biogenesis and cargo loading in C. neoformans, as this will provide further insights into the mechanisms underlying the regulation, production, composition, and diversity of fungal EVs, which would enable a better understanding of their biological function. Expanding our knowledge on pathogenic fungal EVs would pave the way for utilizing native or engineered EVs as promising candidates for therapeutic applications, including fungal infection diagnosis and vaccine development.

Impact of ERQC disruption on glycoprotein folding and EV-mediated transport of virulence factors in C. neoformans

(A) In WT strain, C. neoformans UGGT homolog, Ugg1, functions as a sensor for misfolded glycoproteins within the ER, playing a crucial role in protein quality control. Functional ERQC is essential not only for ensuring the proper folding of glycoproteins, which is critical for maintaining cellular fitness, but also for facilitating EV-mediated secretion of capsule polysaccharides and virulence-related enzymes necessary for pathogenicity. (B) In the UGGT-deficient strain (ugg1Δ), ER stress is increased because of misfolded protein accumulation within the ER lumen. This heightened stress leads to decreased cellular fitness, which negatively impacts EV biogenesis and cargo loading. Consequently, significant defect occurs in EV-mediated transport, which ultimately leads to a complete loss of virulence. Nc: nucleus, Vc: vacuoles

Materials and Methods

Strains, culture conditions, plasmids, and primers

The C. neoformans strains that were constructed and used in this study are listed in Supplementary Table S1A. The plasmids and primers used are listed in Supplementary Tables S1B and S1C, respectively. The construction of the deletion mutants is described in the Supplementary Information. The strains were typically cultured in YPD medium (1% yeast extract, 2% bacto peptone, and 2% glucose) at 30 °C with shaking (220 rpm). C. neoformans transformants were selected by culturing on YPD solid medium containing 100 μg/ml nouseothricin acetyltransferase (Jena Bioscience, Germany; indicated as YPDNAT), YPD solid medium with 100 μg/ml hygromycin B (Sigma-Aldrich, USA; indicated as YPDHyB), or YPD solid medium with 100 μg/ml G418 (Duchefa, Netherlands; indicated as YPDNEO). For capsule induction, C. neoformans cells were cultured in liquid Sabouraud dextrose medium (Difco) at 30 °C for 16 h and incubated in 10% Sabouraud dextrose medium (pH 7.3) supplemented with 50 mM morpholinepropanesulfonic acid (MOPS) at 30 °C for 2 days. For melanin production analysis, the cells were spotted onto an L-DOPA agar plate (7.6 mM L-asparagine monohydrate, 5.6 mM glucose, 22 mM KH2PO4, 1 mM MgSO4·7H2O, 0.5 mM L-DOPA, 0.3 mM thiamine-HCl, and 20 nM biotin) and incubated at 30 °C or 37 °C for 2 days.

HPLC and MALDI-TOF-based N-glycan structure analysis

Cell wall mannoproteins (cwMPs) from C. neoformans were isolated and subjected to N-glycan structure analysis as described previously (Park et al., 2012; Thak et al., 2018). Briefly, N-glycans were released from purified cwMPs through PNGase F treatment (New England Biolabs, UK), followed by purification using a Carbograph Extract-Clean column. The purified N-glycans were labeled with 2-aminobenzoic acid (2-AA; Sigma-Aldrich, USA) and further purified using a Cyano Base cartridge (Agilent, USA). 2-AA-labeled N-glycans were analyzed using a Waters 2690 HPLC system equipped with a 2475 fluorescence detector that was set to excitation and emission wavelengths of 360 nm and 425 nm, respectively. Data were collected using the Empower 2 software (Waters). For MALDI-TOF analysis, neutral N- glycans were collected from HPLC fractionation and dried. The matrix solution was prepared as previously described (Thak et al., 2018) and mixed with the samples in equal volumes. The samples were spotted on an MSP 96 polished-steek target (Bruker Daltonics, Germany), and the crystalized samples were analyzed using a Microflex mass spectrometer (Bruker Daltonics, Germany) in a linear negative mode.

Transmission electron microscopy

C. neoformans cells were cultured either at an initial optical density at 600 nm (OD600) of 0.2 at 30 °C in YPD medium until OD600 reached 0.8 or for 2 days in 10% Sabouraud media at 30 °C. The cell pellets were washed twice in PBS and fixed for 12 h in 2% glutaraldehyde-2% paraformaldehyde in 0.1 M phosphate buffer (pH 7.4). The cells were washed in 0.1 M phosphate buffer and post-fixed with 1% OsO4 in 0.1 M phosphate buffer for 2 h. The cells were dehydrated using an ascending ethanol series (50, 60, 70, 80, 90, 95, 100%) for 10 min each and infiltrated with propylene oxide for 10 min. The specimens were embedded with a Poly/Bed 812 kit (Polysciences Inc., USA) and polymerized in an electron microscope oven (TD-700, DOSAKA, Japan) at 65 °C for 12 h. The block was cut into 200-nm semi-thin sections with a diamond knife in the Ultramichrome and stained with toluidine blue for observation using optical microscopy. The region of interest was further cut into 80-nm thin sections using the ultramicrotome, placed on copper grids, double stained with 3% uranyl acetate for 30 min and 3% lead citrate for 7 min, and imaged using a transmission electron microscope (JEM-1011, JEOL, Tokyo, Japan) equipped with a Megaview III CCD camera (Soft imaging system-Germany) at the acceleration voltage of 80 kV.

Animal study and in vitro survival analysis

Animal studies were conducted at the Chung-Ang University Animal Experiment Center. The study design was approved by the Ministry of Food and Drug Safety (MFDS, South Korea). Survival and fungal burden were assayed as described previously (Cheon et al., 2011). Briefly, eight mice (6-week-old female A/J Slc mice; Japan SLC) per strain were infected with 105 cells via intranasal instillation. The mice were weighed and monitored once daily and euthanized after rapid 30% weight loss or identification of signs of morbidity. Kaplan–Meier survival curves were generated using Prism version 7 (GraphPad Software). For performing the fungal burden assay, the lungs of C. neoformans-infected mice were dissected on days 7 or 60. Half-organ portions of the excised lungs were homogenized, serially diluted, and plated onto YPD medium containing 100 μg/ml chloramphenicol (Sigma-Aldrich, USA). The other half-lung samples were fixed, sectioned, and stained with mucicarmine (Abcam, UK) for histopathological analysis. C. neoformans colonization was analyzed using a Zeiss Axioscope (A1) equipped with an AxioCam MRm digital camera.

Cell survival within macrophages was analyzed by opsonizing C. neoformans cells with 10 mg/ml of 18B7 antibody at 37 °C for 1 h. The macrophage-like J774A.1 (105) cells were seeded onto 96-well plates in DMEM medium supplemented with 10% FBS and cultured at 37 °C in 5% CO2 for 18 h. The opsonized C. neoformans (105) cells were co-incubated with activated macrophages at 37 °C in 5% CO2 for 1 h. The non-phagocytized yeast cells were removed by washing each well thrice with PBS. Then, DMEM medium supplemented with 10% FBS was added to each well, followed by culturing at 37 °C in 5% CO2 for 24 h. The macrophages were lysed in distilled water by vigorous pipetting, and the fungal cells were collected and serially diluted. Cryptococcal survival was assessed using two independent colony forming unit (CFU) assays.

Capsule transfer and shedding analysis

Capsule transfer assays were performed as described previously (Reese and Doering, 2003). Briefly, conditioned medium (CM) was prepared as a source of GXM by culturing the respective strains for 5 days in YPD medium, followed by filtering and storing the culture supernatant at 4 °C. The acceptor strains were cultured overnight at 30 °C in YPD medium. In total, 2 × 106 acceptor cells were incubated with 1 µl of CM for 1 h at room temperature under rotation (18 rpm) and washed twice with PBS. Capsule acquisition was visualized by incubating the cells with an anti-GXM (18B7) antibody conjugated with AlexaFluor 488 (ThermoFisher Scientific, USA) for 1 h at 37 °C and observing under an Eclipse Ti-E fluorescence microscope (Nikon, Japan), equipped with a Nikon DS-Qi2 camera. The images were processed using the NIS-elements microscope imaging Software (Nikon, Japan). Capsule shedding analysis was performed by analyzing GXM shedding using a modified previously described protocol (Yoneda and Doering, 2008). The respective strains were cultured in 10% Sabouraud media for 2 days, and the culture supernatant was sterile filtered. Enzyme denaturation was performed by subjecting the filtrate to heating at 70 °C for 15 min, followed by centrifugation at 13,000 rpm for 3 min. Intracellular GXM analysis was performed by resuspending the pellets in TNE buffer (50 mM Tris-HCl [pH 7.5], 150 mM NaCl, 5 mM EDTA) with the same volume of glass beads (425–600 µm in diameter, Sigma-Aldrich, USA). The cells were disrupted four times for 15 s at 5,000 rpm using a Precellys 24 Tissue Homogenizer (Bertin Technologies, France), followed by centrifugation at 16,000 rpm for 5 min at 4 °C. Next, 10 µl of the supernatant of either secreted or intracellular polysaccharide was mixed with 6X loading dye and run on a 0.6% certified Megabase agarose (Bio-rad, USA) gel in 0.5X TBE (44.5 mM Trisma base, 44.5 mM boric acid, 1 mM EDTA [pH 8.0]) at 25 V for 16 h. The polysaccharides were transferred onto a nylon membrane using the southern blotting technique. The membrane was air-dried, blocked using 5% skim milk, and treated overnight with 2 µg/ml 18B7 antibody. After washing, the membrane was incubated with an anti-mouse peroxidase-conjugated secondary antibody and subjected to detection using chemiluminescence.

Detection of enzymatic activities in intracellular and secretary fractions

Biochemical enzymatic activities of acid phosphatase, urease, and laccase in the intracellular and secreted fractions were assayed spectrophotometrically, using previously described methods (Lev et al., 2014; Fu et al., 2018; de Sousa et al., 2022). Acid phosphatase activity was assayed by culturing cells in MM-KCL medium (0.5% KCl, 15 mM glucose, 10 mM MgSO4.7H20, 13 mM glycine, and 3 µM thiamine) for 3 h at 30 °C. The culture supernatant and soluble cell lysate were allowed to react with 2.5 mM p-nitrophenyl phosphate (pNPP) for 30 min at 37 °C. Urease activity was determined after culturing the cells in Rapid Urea broth (RUH) for 24 h at 30 °C and subjecting the culture supernatant and soluble cell lysate to incubation with phenol red. Laccase activity was assayed by culturing the cells in asparagine salts media (7.6 mM L-asparagine, 0.1% glucose, 22 mM KH2PO4, 1 mM MgSO4.7H2O, 0.3 mM thiamine-HCl, and 20 nM biotin) for 48 h at 30 °C and allowing the supernatant and soluble cell lysate to react overnight with 10 mM L-DOPA. The reactions were quantified by measuring OD420 (acid phosphatase), OD570 (urease), or OD480 (laccase). Cellulase activity was determined by culturing the cells in cellulose media (0.1% NaNO3, 0.1% K2HPO4, 0.1% KCl, 0.5% MgSO4, 0.5% yeast extract, 0.1% glucose, and 0.5% low viscosity carboxymethyl cellulose) for 24 h at 30 °C. The culture supernatants were concentrated using an Amicon tube (30 kDa cutoff, Sigma-Aldrich, USA). The concentrated supernatants and soluble cell lysates were allowed to react at 40 °C for 10 min with the substrate 4,6-O-(3-ketobutylidene)-4-nitrophenyl-β-D-cellopentaoside (BPNPG5), which was provided in the cellulase assay kit (CellG5 Method, Megazyme, Ireland). The reaction was terminated by adding 2% [w/v] Tris buffer (pH 10), and the absorbance of 4-nitrophenol was measured at 400 nm. Cellulase activity (CellG5 Units/ml) was calculated as indicated in the cellulase assay kit. α-amylase activity was measured using the α-Amylase Activity Colorimetric Assay Kit (Biovision Technologies, USA) according to the manufacturer’s instructions. Yeast cells were cultured overnight in YPD medium at 30 °C, and the culture supernatants were concentrated using an Amicon tube (30 kDa cutoff, Sigma-Aldrich, USA). The concentrated supernatants and soluble cell lysates were incubated for 1 h at 25 °C in assay buffer with the substrate ethylidene-pNP-G7, which was provided in the kit. OD405 was measured. All the activity analysis results were normalized according to cell density (OD600).

Purification, nanoparticle tracking analysis (NTA), and Cryo-TEM imaging of extracellular vesicles

EV purification was performed according to a previously published protocol (Reis et al., 2019) with modifications. One loop of cells was inoculated into 10 ml of liquid YPD and incubated at 30 °C for 24 h with shaking (220 rpm). The cells were washed twice with 10 ml of PBS, counted, and diluted in PBS to a density of 3.5 × 107 cells/ml. Aliquots of cell suspension (300 μl) were spread onto synthetic dextrose (SD) solid medium plates and incubated for 24 h at 30 °C. The cells were carefully recovered from each plate using an inoculation loop, gently resuspended in 30 ml PBS, and pelleted through centrifugation at 4,000 rpm for 10 min at 4 °C. The supernatant was collected and centrifuged again at 15,000 ×g for 15 min at 4 °C. Then, the supernatant was filtered through 0.45-μm syringe filters and ultracentrifuged at 100,000 ×g for 1 h at 4 °C (MLA-50 fixed angle rotor, Beckman Coulter, Germany). The supernatant was discarded, and the EV pellets were collected and resuspended in 1 ml of PBS for immediate use or stored at −80 °C for further experiments. The samples were diluted 100-fold, and EV sizes were measured using an NTA instrument (Nanosight Pro, Malvern Panalytical, Netherlands) coupled to a 532-nm laser (Malvern Panalytical, Netherlands), SCMOS camera (Hamamatsu Photonics. Japan), and syringe pump (Malvern Panalytical, Netherlands). The data were analyzed using the NS Xplorer software (v1.1.0.6, Malvern Panalytical, Netherlands).

Cryo-TEM imaging was performed by loading purified EVs (3 µl) onto a lacey carbon grid (Lacey Carbon, 300mesh Cu, Ted Pella Inc., USA), which was glow discharged at 15 mA for 60 s. The sample-loaded grid was blotted for 3 s at 15 °C and 100% humidity and immediately plunge-frozen in liquid ethane. The process was performed by Vitrobot Maek IV (Thermofisher Scientific, USA, SNU, CMCI). The frozen grids were imaged using TEM (JEM-2100F, JEOL, Japan); the temperature of the grid was maintained at approximately −180 °C at an acceleration voltage of 200 keV. The images were recorded using an ultrascan 1000 electron detector.

EV proteome analysis

EVs proteins were solubilized in 8 M urea, 100 mM Tris (pH 7.5), and 5 mM tris (2-carboxyethyl) phosphine (TCEP) for 20 min at 23 °C. The protein concentration was quantified using the Protein Assay Dye (Bio-rad, USA). A Protifi S-TrapTM mini spin column (C02-mini-80, Protifi, USA) was used to process each sample. After digesting with trypsin gold (V5280, Promega, USA), the digested peptides were eluted using 50 mM tetraethylammonium bromide (TEAB; Thermo Fisher Scientific, USA), 0.2% formic acid, and 50% acetonitrile. The pooled peptide solution was dried, dissolved in 100 mM TEAB, and labeled using TMTpro™ 16plex Label Reagent Set (Thermo Fisher Scientific, USA). The labeled peptides (total, 100 µg proteins) were combined prior to offline basic reverse-phase liquid chromatographic (bRPLC) fractionation. Linear gradient was performed using buffer A (10 mM TEAB in water) and buffer B (10 mM TEAB in 90% acetonitrile), and 10 fractions were analyzed in total using am LC-MS/MS system. The samples were dissolved in 0.1% formic acid using an UltiMate 3000 RSLCnano system and analyzed using an Orbitrap Eclipse Tribrid mass spectrometer (Thermo Fisher Scientific, USA). All MS raw files were converted into mzML and ms2 file formats using the MSConvert (version 3.0.20033) software.

The C. neoformans proteome was determined by downloading a protein FASTA file from Uniprot (http://uniprot.org), which included 30,647 reviewed (Swiss-Prot) and unreviewed (TrEMBL) proteins entries. A proteome search database with reversed sequence and contaminants in the Integrated Proteomics Pipeline (IP2, version 5.1.2, Integrated Proteomics Applications Inc., San Diego, CA) was generated. The proteome search from 20 ms2 files was performed with IP2 and its following parameters, followed by evaluation for a false discovery rate (FDR) using IP2 and Proteininferencer (version 1.0, Integrated Proteomics Applications Inc., USA). Protein quantification and statistical analysis for discovery of differentially expressed proteins (DEPs) was performed using the ms2 files with tandem mass tag (TMT) reporter ions using an in-house program coded using Python 3.8, where t-test and Pearson’s correlation analysis between comparison samples was performed using the scikit-learn (version 0.23.2), Scipy (version 1.6.0), and statsmodels (0.12.1) Python libraries.

eLife Assessment

This important study confirms the molecular function of putative components of the N-glycan-dependent endoplasmic reticulum protein quality control (ERQC) system in the pathogen Cryptococcus neoformans. The study demonstrates an involvement in fitness, virulence, and the secretion and composition of extracellular vesicles, albeit in ways that are not yet fully understood. The evidence provided is largely convincing, with rigorous well-controlled assays and the use of complemented strains.

Reviewer #1 (Public review):

Summary:

Using gene deletion analysis, the authors confirm the molecular function of putative components of an N-glycan-dependent endoplasmic reticulum protein quality control (ERQC) system (UGG1, MNS1, MNS101, MNL1, and MNL2), in the basidiomycetous fungal pathogen Cryptococcus neoformans. Specifically, they confirm the essential role of these components in the ERQC system and their role in ER stress which contributes to cellular fitness and pathogenicity.

The second part of the study links the components to secretion, mainly EV biogenesis and composition. However, this part of the study is less convincing.

Strengths:

Although it is unclear why ER stress in the mutants would not manifest into a classical secretion defect, this is a rigorous, well-controlled study, with the use of complemented strains that demonstrate phenotypic restoration. The diagram in Figure 1 is very useful in orientating the reader to a complex subject matter, although the legend could be more descriptive.

Weaknesses:

A major weakness is the sheer volume of data presented (in the main text and supplement), which makes the results difficult to follow and retain: the work could essentially be two separate studies.

Another major weakness is the lack of mechanistic insight into the role of the ERQC system in EV secretion and its disconnection to "classical" secretion, which is difficult to reconcile. Some insight into why EV secretion is decreased, and classical secretion is unaffected, would strengthen the significance of the findings. No mechanism is provided to explain why the ERQC mutants (Ugg1 mutant in particular) would have reduced and heterogeneously sized EVs. Furthermore, it is not convincing that the EV content changes would greatly impact fitness and virulence. The proteomics data showing reduced cargo in the Ugg1 mutant is not convincing and difficult to follow.

Reviewer #2 (Public review):

Summary:

This study investigates the molecular function of the N-glycan-dependent endoplasmic reticulum protein quality control system (ERQC) in Cryptococcus neoformans and correlates this pathway with key features of C. neoformans virulence, especially those mediated by extracellular vesicle transport. The findings provide valuable insights into the connection between this pathway and the biogenesis of C. neoformans extracellular vesicles.

Strengths:

The strength of this study lies primarily in the careful selection of appropriate and current methodologies, which provide a solid foundation for the authors' results and conclusions across all presented data. All experiments are supported by well-designed and established controls in the study of C. neoformans, further strengthening the validity of the results and conclusions drawn from them. The study presents novel data on this important pathway in C. neoformans, establishing its connection with C. neoformans virulence. Interestingly, the findings led the authors to understand the relationship between this pathway and the transport of key fungal virulence factors via extracellular vesicles. This was demonstrated in the study, paving the way for a deeper understanding of extracellular vesicle biogenesis-a field still filled with gaps but one to which this study contributes solid data, helping to clarify aspects of this process.

Weaknesses:

I do not see significant weaknesses in this study. The experiments are well-grounded, and the results are clearly presented. I believe the only weakness is that the paper could be condensed. Sections like the discussion, for instance, are extremely lengthy, which may make reading and, consequently, understanding more challenging for many readers. Regarding the presentation of the results, while clear, the figures contain a lot of information, and I believe that some of this content could be moved to supplementary figures.

Reviewer #3 (Public review):

Summary:

Cryptococcus neoformans is a global critical threat pathogen and the manuscript by Mota et al demonstrates that the pathogen's N-glycan-dependent protein quality control system regulates the capacity of the fungus to cause disease. The system makes sure that glycoproteins are folded correctly. The system is involved in the fitness and virulence of the fungus by regulating aspects of cellular robustness and the trafficking of virulence-associated compounds outside of the cell via transport in extracellular vesicles.

Strengths:

The investigators use multiple modalities to demonstrate that the system is involved in cryptococcal pathogenesis. The investigators generated mutant C. neoformans to explore the role of genes involved in the protein folding system. Basic microbiology, genetic analyses, proteomics, fluorescence and transmission microscopy, nanotracking analyses, and murine studies were performed. The validity of the findings are thus very high. Hypotheses are robustly demonstrated.

Weaknesses:

Aspects of the results should be better explained. Some results are extrapolated in their meaning beyond the extent of the data.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Recommendations for the authors):

Major comments

(1) The section on page 20 describing the proteomic analysis of EVs is poorly written and confusing, with a lot of data in the supplement. It is not clear what the proteomics data actually means.

We appreciate your feedback on the clarity of the proteomic analysis section. We have rewritten the section on page 20 with more detained information to provide a clearer explanation of the proteomics data and its biological significance. Additionally, we have incorporated a comparative analysis of the EV and total cell lysate proteomes (Fig. 8E, Supplementary Fig. S7A, Supplementary Tables 3 and 4) for supplemental data interpretation.

(2) The order of the data could be improved.

We appreciate your feedback regarding the data organization. We have reorganized the order and position of some data in a more structured and coherent manner, as suggested by the reviewers.

- Reorganization of the qPCR data (previously Fig. 1C) as Fig. 3A

- Removal of the data on the growth analysis on raffinose media (previously Fig. 7H).

-Reorganization of the spotting data of the double mutant (previously Fig 3B) to Supplementary Fig. S3B

- Reorganization of the subcellular localization data (previously Fig 3E) to Supplementary Fig. S4A

(3) The discussion is repetitive with the introduction and merely summarizes the results and speculates on the mechanism of how the absence of UGGT, leading to ERQC defects, results in defective EV biogenesis/cargo loading in C. neoformans.

We removed several repetitive sentences in the discussion and provided additional information on proteome analysis.

Other questions and comments

(1) Instead of comprehensively analyzing EVs from the UGG1 mutant, a more informative approach to better understanding how defects in N-linked glycosylation impact secretion, would be to do a proteomic analysis on the total secretions (including beta glucanase-treated cells to release classically secreted proteins from the cell wall) and EVs.

We agree that a comprehensive proteomic analysis of total secretions and classically secreted proteins would provide deeper insights into how defects in N-glycosylation impact secretion in C. neoformans. To address this concern, we performed an additional set of proteomic analyses, the proteome profiles of total cell lysates and the secretome of C. neoformans cultivated in SD broth and presented the results as Supplementary Table S5 and Supplementary Fig. S7B. These additional analyses provide further insights into the impact of UGG1 deletion on both conventional and unconventional secretion pathways, supporting a more pronounced effect of the UGG1 defect on EV-mediated trafficking. The discussion has been updated accordingly (Page 22, lines 509-514).

(2) The melanization defect in Ugg1 mutant is not strong. Could the reduction be due to partially compromised Ugg1 mutant growth at 30{degree sign}C as indicated in the spot tests. Were photos of the spot dilution assays taken at 1 and 2 days to investigate slower growth? Or alternatively were growth curves taken in a liquid culture?

For accuracy of melanin synthesis defect, in addition to analysis on L-DOPA plates, we had assessed melanin production in liquid L-DOPA medium following a 3-day incubation, and the melanin production in liquid media was normalized by cell density (OD600). The data on normalized melanin production is now included as Fig. 4B in the revised manuscript. The defective laccase activity in the ugg1Δ mutant (Fig. 7C) further corroborates our melanization assay results, which is additionally mentioned in the text (Page 18, lines 393-395).

(3) Is it accurate to say that some virulence factors (i.e. melanin, capsule and phosphatases) are predominantly trafficked through EV's in C. neoformans? Have studies been done to determine the proportion of virulence factors trafficked via EV's versus traditional secretion?

We thank you for the thoughtful comments. Some virulence factors, such as urease, melanin and capsule polysaccharides, lack a signal peptide required for targeting for the conventional ER/Golgi secretion pathway. It is generally assumed that the trafficking of these factors in C. neoformans is predominantly mediated by non-conventional secretion via EVs. Additionally, even some virulence factors with signal peptides, such as laccase and phosphatases, are also transported via EVs besides the conventional secretion. The quantitative analysis to compare the proportion of virulence factors secretion via EVs versus the conventional pathway has not been yet reported, despite that genetic evidence suggests that conventional secretion also plays a significant role in the export of capsule polysaccharides. Thus, we were also careful not to highlight EV as the main route of virulence factors in the manuscript.

(4) There is insufficient background in the introduction linking what is known about the ERQC process to secretion in general. The topic changes from the ERQC process to fungal virulence factor, with a primary focus on non-classical (EV-based) secretion. Classical secretion should also be discussed without assuming that non classical (EV) secretion is the major pathway contributing to fungal virulence.

We appreciate your insightful comments highlighting the need for more background on the ERQC process and its relationship with secretion. To address the reviewer’s concerns, we have added sentences to describe the key roles of ERQC in conventional protein secretion in the Introduction (Page 5, lines 102-106).

(5) Figure 1A. What does the blue filled circle with the red outline signify? Fig 1 A legend is not well explained. A summary using material provided in the intro/discussion should be included to briefly explain the process and the differences between fungal species. Please also be aware that the intro starts describing the human ERQC process and then switches to what happens in S. cerevisiae.

We have revised Figure 1A by removing the red circle and updated the figure legend in the revised manuscript to include more detailed information about the ERQC differences across higher eukaryotes and fungal species.

(6) Figure 2A. There are no units on the Y-axis. Presumably, the scale is the same for all 3 strains.

Thank you for your comments. The Y-axis is the same for all three strains and, as in Fig. 2C, and represents the relative fluorescence intensity obtained from the HPLC analysis. We added the units on the Y-axis in Fig. 2A.

(7) If Mnl1 and 2 have proposed roles in proteasomal degradation, wouldn't they be expected to have ER retention signals, like Ugg1?

We appreciate your valuable insights regarding the absence of ER retention signals in Mnl1 and Mnl2. Previous studies have shown that Saccharomyces cerevisiae Mnl1/Htm1 does not possess canonical KDEL/HDEL-like ER retention signals. Instead, its retention in the ER lumen is facilitated through its interaction with protein disulfide isomerase Pdi1, which contains an HDEL sequence (Gauss et al. 2011). Thus, it is expected that non-canonical retention mechanisms—such as interactions with other ER proteins—could contribute to the retention of Mnl1 and Mnl2 within the ER. We added this information to the revised manuscript (Page 8, lines 154-159).

(8) Figure 1 C qPCR showing change in mRNA in response to ER stress should not be grouped in this figure. It could be standalone or discussed when the spot dilution assays are performed. Anyway, spots tests are more convincing of a role in stress response than qPCR as the ugg1 mutant is sensitive to tunicamycin, DTT and cell wall stressing agents.

As suggested by the reviewer, we have reorganized the qPCR data as a part of Figure 3 (Figure 3A) in the revised manuscript.

(9) It is odd that mns1/101 mutants are not sensitive to ER and CW stress given their proposed differing location/function in the pathway (Figure 1) determined from the N-linked profiling. Any explanation? Could there be redundancy?

We appreciate the reviewer’s observation regarding the lack of ER and CW stress sensitivity in the mns1Δ and mns101Δ mutants, despite their proposed roles in N-glycan processing. We had previously reported that the C. neoformans alg3Δ mutant, lacking a critical enzyme responsible for the synthesis of Dol-PP-Man6GlcNAc2 in the N-glycosylation pathway, exhibited clearly impaired N-glycan elongation, but showed no detectable growth defects even under stress conditions in vitro. However, alg3Δ is avirulent in in vivo pathogenicity (Thak et al., 2020). Similarly, the mns1Δ101Δ double mutant shows glycan-processing defects that do not compromise cellular fitness under stress conditions but result in attenuated virulence in animal models. These findings suggest that some glycosylation-related defects may impact more severely in vivo pathogenicity rather than in vitro stress sensitivity.

(10) Although the Silver-stained gels of the ugg1 mutant are not particularly informative, why weren't they (and Con A blots) performed for the other mutants?

The overall decrease of hypermannosylated glycans observed in the ugg1Δ mutant allowed us to detect clear alterations in protein glycosylation patterns in the lectin blot using Galanthus nivalis agglutinin, which recognizes terminal α1,2-, α1,3-, and α1,6-linked mannose residues. In contrast, the limited changes of a few glycan species in other mutants, including mns1Δ, mns101Δ, and mns1Δ101Δ, are relatively subtle to be detected in the lectin blot, due to only minor differences in the average lengths of their N-glycans compared to the WT. Therefore, we presented the lectin blotting data only for the ugg1Δ mutant.

(11) If there is ER stress under normal conditions in the Ugg1 mutant then technically this mutant should be growing more slowly under normal conditions. This is difficult to predict in a spot dilution assay where growth is only visualized at day three when any growth defect may have been corrected. The slower growth rather than the reduced secretion of GXM specifically is therefore more likely to be responsible for the reduced virulence.

We appreciate the reviewer’s insightful comment regarding the interplay between ER stress, growth defects, and virulence attenuation in the ugg1Δ mutant. While retarded growth in C. neoformans is often associated with reduced virulence, there are a few exceptions. For instance, disruptions in cell cycle progression in C. neoformans have been reported to result in larger capsule sizes, which rather enhance in vivo virulence when analyzed in Galleria mellonella infection models (García-Rodas et al., 2014). This highlights that growth defect alone is not sufficient for virulence attenuation. In the case of the ugg1Δ mutant, we speculate that the almost complete loss of virulence is attributed not only to its growth retardation but also to its impaired secretion of key virulence factors, including the polysaccharide capsule.

(12) The rationale for using leucine analogue 5',5',5'-trifluoroleucine (TFL), in a growth assay (Fig. 3C) to determine whether the defective ugg1Δ phenotypes are induced by ER stress caused by misfolded protein accumulation is not explained.

The leucine analogue 5',5',5'-trifluoroleucine (TFL) can be incorporated into newly synthesized proteins, disrupting normal folding and thus leading to the generation of misfolded proteins (Trotter et al., 2002; Cowie et al., 1959). In the context of a defective ERQC pathway, these misfolded proteins cannot be adequately repaired, resulting in their accumulation and triggering ER stress. Excessive ER stress may ultimately inhibit cell growth in the presence of TFL. This explanation has been incorporated into the revised manuscript (Page 11, lines 236–241).

(13) I would argue that only the Ugg1 and double Mns mutant were defective in virulence. For the single mutants, it looks like no difference was found relative to WT. The longer median survival of these mutants (if significant) is most likely due to poor infection technique.

We agree with the reviewer’s opinion that the mns1Δ and mns101Δ single mutants have no significant difference in in vivo virulence compared to the WT strain, unlike the mns1Δ101Δ double mutant which showed significant attenuated virulence. We had previously addressed that in the manuscript (Page 13, lines 267-269).

(14) The authors conclude that the ugg1Δ strain specifically is impaired in extracellular secretion of capsular polysaccharides but is this via classical (SAV1) secretion or EVs?

In addition to EV-mediated transport, capsular polysaccharide secretion can occur via the Sav1 (Sec4p)-mediated classical secretion pathway. However, our proteome data of total cell lysates indicated that the protein levels of Sav1 were comparable between the WT and ugg1Δ strains, suggesting that Sav1p function itself might not be impaired. Given that the ugg1Δ mutant exhibits altered vesicular structures (Supplementary Fig. S6) and loss of microvesicles (Fig. 8A), we speculate that a defect might occur at a post-Sav1p step, such as vesicle fusion with the plasma membrane, likely contributing to the complete defect in secretion of capsular polysaccharides in the ugg1Δ strain, in which EV biogenesis and defective cargo loading are severely impaired, producing EVs that lack capsular polysaccharides (Figure 8F). However, further studies should be carried out to define the contribution of SAV1 to the secretion of capsular polysaccharides in in the ugg1Δ strain.

(15) The rationale for doing 7 H is very confusing.

The experiment assessing raffinose utilization as a carbon source was inspired by the previous work of Garcia-Rivera et al., reporting that the cap59Δ mutant is unable to utilize raffinose due to a defect in the secretion of raffinose-hydrolyzing enzymes. As another way to investigate potential defects in the conventional secretion pathway, we investigated the growth of the ugg1Δ mutant in the presence of raffinose. Due to our extensive data length, we have decided to remove this complementary data from the manuscript.

(16) It is speculated in the discussion that ER stress impacts lipid/sterol synthesis and that LDs (lipid droplets?) aid the UPR and ERAD in degrading misfolded proteins during ER stress in S. cerevisiae. The authors mention that they observed a drastic increase in LDs in the ugg1Δ mutant. Where is this data? Even with the data, this is all speculation. The authors also speculate that increased numbers of vacuoles in ugg1 (where is the data?) could be the cause of the altered vesicular structures observed in the mutants, which may indicate abnormal lipid homeostasis caused by the ERQC defects, which could, in turn, affect EV biogenesis. Again, this is speculative.

The data on lipid droplets (LDs) and vacuole staining are presented in Supplementary Figure S6, showing a drastic increase in LDs and an increased in vacuolar size in the ugg1Δ mutant compared to the wild-type strain, especially in capsule-inducing conditions. In addition to such changes in vesicular structures, our preliminary data on sphingolipids and sterol analysis in the surface lipid fraction of the ugg1Δ mutant led us to propose the hypothesis that ERQC defects may impact lipid metabolism, which in turn could influence EV biogenesis and membrane properties. It is expected that these findings would provide a strong foundation for future studies exploring the link between ERQC, lipid homeostasis, and EV biogenesis. We have revised our speculation on the association of abnormal lipid homeostasis, caused by ERQC, with EV biogenesis more appropriately by adding the information on our preliminary data of lipid profiles and mentioning that the ugg1Δ mutant lacks microvesicles, which are derived from the plasma membrane (Page 24, lines 554-559).

Reviewer #2 (Recommendations for the authors):

(1) My suggestions for the authors are the same as those presented in the public review: (1) reducing the text in certain sections of the paper to improve readability for the audience, and (2) reconsidering the figures to reduce the amount of information in each one, moving some of the content to the supplementary material.

We thank the reviewer for their constructive suggestions regarding the organization and readability of the manuscript. As suggested, we addressed your concerns as follows:

(1) Reducing the text in the Introduction, Results, and Discussion sections by removing repetitive statements and simplifying complex descriptions where possible.

(2) Changing the presentation of figures: we have also reorganized the presentation of some data by moving non-essential data to the supplementary material. The updated figures and supplementary materials have been clearly referenced in the text to guide readers.

(3) Reorganization of materials and methods: some parts of methods were moved to Supplementary Information

(4) Removal of Figure 7H and the sentences describing the result

More detailed explanations on the reduction and reorganization are also described in the response to the major comments (2) and (3) made by Reviewer #1.

(2) Figure 3, for example, shows no difference in fungal growth under different cultivation conditions. This information is valuable but could be mentioned in the text, with the image provided as supplementary material, focusing the figure only on images that show significant growth differences among the strains. I suggest a similar approach for other figures so that the authors can include only the most relevant results in the main body of the article and move some figures to the supplementary materials.

For Fig. 3, the spotting data of the double mutant (previously Fig. 3B) is now presented in the supplementary information (Supplementary Fig. S3B). Additionally, the subcellular localization data (previously Fig 3E) was also moved to the supplementary material (Supplementary Fig. S4A).

Reviewer #3 (Recommendations for the authors):

(1) Line 43 "EV-mediated transport of virulence bags" doesn't make sense. EVs have been described as "virulence bags" (and are in this work later in the introduction) but this should here be "transport of virulence factors" or "compounds associated with virulence" but only if you have confirmed that the "cargo" is consistent with this- which is not evident in the abstract.

Thank you for your insightful comment. We have revised this to "EV-mediated transport of virulence factors" in line with your suggestion.

(2) Line 49 "secretory pathway" - is there not more than one secretion pathway?

Thank you for pointing this out. The term "secretory pathway" has been updated to "secretory pathways" to acknowledge the presence of both conventional and unconventional secretion mechanisms.

(3) Line 53 "recognizes folding defects, repairs them, and ensures the translocation of irreparable misfolded proteins" should be "recognizes folding defects and repairs them or ensures the translocation of irreparable misfolded proteins.

Thank you for pointing this out. We have revised the sentence as you suggested.

(4) Lines 88-90 ALG needs to be written out the first time - Asn-linked glycans. Also, consider adding that ALG genes are present in most eukaryotes as it is unclear what you are comparing C. neoformans to.

Thank you for your helpful comment. We have revised the text to write out "ALG" as "Asn-linked glycosylation" and added the sentence “ALG genes are evolutionary conserved in most eukaryotes” in the revised manuscript (Page 4, line 84).

(5) Line 99 Cryptococcus has already been abbreviated to C. so don't write it out again.

We have corrected "Cryptococcus" to “C.” throughout the manuscript after its first mention.

(6) Line 152- tunicamycin and DTT are not described yet, which may make it challenging for some readers to understand what these drugs are doing/why they were used. What is on lines 156 and 157 for these drugs should go up with the first mention of these drugs.

Thank you for your helpful suggestion. We have revised the manuscript to include the descriptions and purpose of using tunicamycin (TM) and dithiothreitol (DTT) immediately following their first mention, as recommended (Page 10, lines 208-210).

(7) The text for Figure 1 C is inaccurate. High temperature also induced KAR2, as noted above, but inaccurately stated in line 160. There is no comment on the significant UGG1 increase with tunicamycin or that KAR2 was highest in this condition.

Thank you for your thoughtful comment. We have better clarified the significant increase of UGG1 expression following tunicamycin treatment and KAR2 induction upon heat stress in the revised manuscript (Page 10, lines 216-217). Please note that Fig. 1C was revised and is now referred to as Fig. 3A.

(8) Figure 2B is not well explored/explained. There appears to be more protein in the mutant, including of higher weight in the intracellular compartment. It is difficult to ascertain if there is more too in the secretion phase with this gel. The methods do not specifically describe the concentration of protein added - just volume. Is what we are seeing a loading issue vs real differences?

Thank you for your insightful comments regarding Figure 2B. We added information on amounts of protein (30 µg per lane) in the legend of Figure 2B.

The main purpose of Fig. 2B is to examine the altered glycosylation pattern of ERQC by detecting glycoproteins using the Galanthus nivalis agglutinin, which specifically bind terminal α1,2-, α1,3-, and α1,6-linked mannose residues. The result of lectin blotting indicated that glycoproteins are more abundantly detected in the secretion fraction compared to in the soluble intracellular fraction, consistent with the general notion that more than 50% of secretory proteins are glycoproteins. Also, the more abundant proteins with decreased molecular weight in the secretion fraction of ugg1Δ mutant supported the N-glycan profiles with decreased hypermannosylation in ugg1Δ mutant. We added the purpose and more detailed interpretation on Figure 2B in the revised manuscript (Page 9, lines 174-179).

(9) Line 242 "melanin pigment" is redundant as melanin is a pigment.

We thank the reviewer for pointing out the redundancy in the phrase. We revised the text to simply state "melanin".

(10) Line 250 drops "completely" especially as the mutant did colonize the lungs of mice.

To avoid any possible misleading, we removed the term "completely" in the revised manuscript.

(11) Line 275- need to reference 18B7 as it is first introduced here.

We added the reference on the antibody 18B7 in the revised manuscript.

(12) Line 308- there are specific techniques to measure GXM size that could validate or refute the statement on "incomplete" polysaccharides. For example, DOI:10.1128/EC.00268-09.

We appreciated the valuable suggestion on specific techniques to measure GXM size, which will be one of key experiments in our future study. In the revised manuscript we cited the suggested reference to indicate the need for validation of our statement (Page 14, lines 316-318).

(13) Line 496 "mammals" - why is this used when the study is on a fungus, not a mammal? The structure of the first 2 paragraphs can be clearer to focus more on fungal biology.

We have compared both mammals and fungi to emphasize that the ERQC system is conserved among eukaryotes but diverged with a few species-specific features. This comparison is relevant in the context of understanding the evolutionary unique features of ERQC pathways in C. neoformans. We modified the first 2 paragraphs to clarify the main issue of our present study (Page 21, lines 472-483).

(14) Line 525- the ugg mutant was not avirulent as CFU was present and histopathology in the supplementary figures shows the tissue with ugg1 deletion was not normal (although the images are not especially easy to review). Yes, the mutant did not kill under your test conditions, but it was not avirulent (incapable of causing disease). Significantly attenuated or other descriptors should be utilized. Line 548 is also thus incorrect "complete loss of virulence").

We appreciate the reviewer’s concern regarding the description of the ugg1Δ mutant as avirulent. We agree that the use of merely “avirulent" may not fully capture the observed phenotypes in the CFU and histopathological data, since we cannot exclude the possibility that the ugg1Δ mutant retains the ability to establish an infection. Thus, we have revised the text by describing the ugg1Δ mutant as "almost avirulent".

(15) Line 597- the study by Fukuoka used kidney cells. It is misleading to not clearly state that this finding of ER stress was NOT done in fungi as the way it is presented makes it read as if this work was performed in C. neoformans. This should be clarified. This should also be double-checked and clarified for other statements, such as the reference to Harada in line 606, as this study used melanoma cells. These cell types are very different from cryptococcus- though I absolutely concur that lessons can be learned from comparative assessments.

We thank the reviewer for pointing out the need to clarify the experimental context of the cited studies. We explicitly stated the host cell types used in the referenced studies by Fukuoka et al. and by Harada et al., respectively, in the revised manuscript (Page 25, lines 560 and 568).

Version 2

Abstract

Abstract

A conserved N-glycan-dependent endoplasmic reticulum protein quality control (ERQC) system has evolved in eukaryotes to ensure accuracy during glycoprotein folding. The human pathogen Cryptococcus neoformans possesses a unique N-glycosylation pathway that affects microbial physiology and interactions with the infected host. To investigate the molecular features and functions of the ERQC system in C. neoformans, we characterized a set of mutants with deletion of genes coding for the ERQC sensor UDP-glucose:glycoprotein glucosyltransferase (UGG1) and putative α1,2-mannose-trimming enzymes (MNS1, MNS101, MNL1, and MNL2). The ugg1Δ, mns1Δ, mns101Δ, and mns1Δ101Δ mutants showed alterations in N-glycan profiles, defective cell surface organization, decreased survival in host cells, and varying degrees of reduced in vivo virulence. The ugg1Δ strain exhibited severely impaired extracellular secretion of capsular polysaccharides and virulence-related enzymes. Comparative transcriptome analysis showed the upregulation of protein folding, proteolysis, and cell wall remodeling genes, indicative of induced ER stress. However, no apparent changes were observed in the expression of genes involved in protein secretion or capsule biosynthesis. Additionally, extracellular vesicle (EV) analysis combined with proteomic analysis showed significant alterations in the number, size distribution, and cargo composition of EVs in ugg1Δ. These findings highlight the essential role of the functional ERQC system for cellular fitness under adverse conditions and proper EV-mediated transport of virulence factors, which are crucial for the full fungal pathogenicity of C. neoformans.

Introduction

Glycoproteins, destined for the secretory pathways, enter the endoplasmic reticulum (ER) lumen where protein folding occurs (Rapoport, 2007). The accumulation of misfolded proteins affects cell viability and homeostasis; therefore, eukaryotes evolved a conserved ER quality control (ERQC) system that recognizes folding defects, repairs them, or ensures the translocation of irreparable misfolded proteins into the cytosol for proteasome-mediated degradation via the ER-associated degradation (ERAD) system (Thibault and Ng, 2012; Xu and Ng, 2015; Balchin et al., 2016), a process that heavily relies on N-glycosylation to determine protein folding conformation (Aebi, 2013; Varki, 2017).

Many aspects of N-glycosylation are highly conserved, and most eukaryotes initiate this process by synthesizing a Dol-PP-linked Glc3Man9GlcNAc2 oligosaccharide as a common core N-glycan, which attaches to nascent polypeptides in the ER via an asparagine residue. Glucosidases I and II (Gls1 and Gls2) remove two glucose residues from the core oligosaccharide. Proteins containing monoglucosylated N-glycans (Glc1Man9GlcNAc2) enter a calnexin and/or calreticulin (CNX/CRT in mammals; Cne in yeast) chaperone-mediated folding cycle. Finally, Gls2 cleaves the remaining glucose residue. If proteins are misfolded, they are recognized by the ERQC checkpoint enzyme UDP-glucose:glycoprotein glucosyltransferase (UGGT), which reglucosylates them for re-entry into the folding cycle (Fig. 1A). Then, the N-glycans of the accurately folded proteins are further processed by Gls2 and α1,2-mannosidase I (Mns1) and moved to the Golgi apparatus. However, irreparably misfolded glycoproteins are targeted for ERAD, where they undergo demannosylation and retro-translocation for proteasomal degradation in the cytosol. Recently, N-glycan precursors in the ER of some protists and fungi were found to be shorter than the typical 14-sugar N-glycan precursors in most eukaryote organisms. The length of these N-glycan precursors significantly impacts N-glycan-dependent QC of glycoprotein folding and ERAD (Banerjee et al., 2007; Samuelson et al., 2015).

Presence of UDP-glucose:glycoprotein glucosyltransferase (UGGT) and α1,2-mannosidases as endoplasmic reticulum protein quality control (ERQC) components in Cryptococcus neoformans.

(A) Schematic representation of the ERQC pathway in Homo sapiens, Saccharomyces cerevisiae, and Cryptococcus neoformans. In mammals, the oligosaccharyltransferase (OST) complex attaches Glc3Man9GlcNAc2 to nascent polypeptides, followed by glucose trimming by glucosidases I and II (GLS1/GLS2). This generates Glc1Man9GlcNAc2, which binds to calnexin (CNX) or calreticulin (CRT) for folding. UGGT reglucosylates misfolded proteins, allowing refolding, while properly folded proteins undergo mannose trimming by ERManI before Golgi transport. In contrast, fungal ERQC systems differ in key components. S. cerevisiae lacks UGGT, relying instead on Gls1/Gls2 and calnexin (Cne1). C. neoformans possesses UGGT but lacks ER glucosyltransferases (Alg6, Alg8 and Alg19) and CRT, resulting in a distinct ERQC system. Domain structures of proteins encoded by C. neoformans UGG1 (CNAG_03648), MNS1 (CNAG_02081), MNS101 (CNAG_03240), MNL1 (CNAG_01987), and MNL2 (CNAG_04498).

The basidiomycetous fungus Cryptococcus neoformans is an opportunistic encapsulated human pathogen that primarily affects immunocompromised individuals, causing fatal meningoencephalitis (Gottfredsson and Perfect, 2000; Kwon-Chung et al., 2000). The N-glycosylation pathway of C. neoformans is evolutionarily conserved; nevertheless, the structure and biosynthesis of its N-glycans includes several unique features (Park et al., 2012). C. neoformans contains serotype-specific high-mannose-type N-glycans with or without a β-1,2-xylose residue attached to the trimannosyl core. Additionally, acidic N-glycans of C. neoformans contain xylose phosphates attached to the mannose residues both within the N-glycan core and outer mannose chains. The intact core N-glycan structure is crucial for C. neoformans pathogenicity (Thak et al., 2020); hence, alterations in the N-glycan structure modulates the interaction between the cell surface mannoproteins and host cells (Lee et al., 2023). Additionally, C. neoformans lacks homologous genes to the Asn-Linked Glycosylation (ALG) genes ALG6, ALG8, and ALG10, which are evolutionary conserved across most eukaryotic organisms and encode the glucosyltransferases that add the glucose residues to the core N-glycan before its attachment to proteins (Park et al., 2012). Man7GlcNAc2 and Man8GlcNAc2 without glucose residues are primarily detected in Dol-PP-linked glycans of C. neoformans (Samuelson et al., 2005). Moreover, the mature core N-glycan structures assembled on the cell surface mannoproteins of C. neoformans are primarily Man6–7GlcNAc2, which are shorter than the expected Man8GlcNac2 (Park et al., 2012). This observation led to the speculation that the terminal α-1,2-mannose residues of C. neoformans N-linked glycans may be more susceptible to trimming by ER α-1,2 mannosidases due to the lack of glucose residues compared to most eukaryotes. Alternatively, the presence of multiple α-1,2 mannosidases may generate more extensively trimmed core N-glycans in C. neoformans.

C. neoformans employs several virulence factors, including an extensive polysaccharide capsule composed of glucuronoxylomannan (GXM) and galactoxylomannan (GalXM) (Doering et al., 2009), melanin (Qiu et al., 2012), and various extracellular enzymes such as phosphatase and urease (Singh et al., 2013), all of which contribute to immune evasion and enhance fungal pathogenicity. To facilitate the export of these virulence-associated molecules, C. neoformans employs both conventional (Yoneda and Doering, 2006; Panepinto et al., 2009) and unconventional secretion pathways (Rodrigues et al., 2007; Casadevall et al., 2019; Rizzo et al., 2021). Classical secretion relies on the ER-Golgi network, while non-classical mechanisms, such as extracellular vesicle (EV)-mediated transport, provide an alternative route for cargo delivery beyond the conventional secretory pathway. C. neoformans EVs have been described as a heterogeneous population of “virulence bags” containing numerous fungal survival and pathogenicity-associated cargo. Proteomic analyses of EV cargo have identified several cell surface glycoproteins, including members of the CDA family, well-known immunomodulators (Specht et al., 2017; Rizzo et al., 2021). The ERQC system is essential for maintaining protein homeostasis by ensuring proper folding, glycosylation, and directed trafficking of secreted proteins to downstream compartments within the conventional secretory pathway. Defects in ERQC affect the fidelity of protein secretion not only by accumulation or degradation of misfolded proteins, leading to the impaired secretion of functional proteins, but also by the escape of misfolded forms, thus resulting in increased secretion of non-properly processed proteins (Marcus and Perlmutter, 2000; Chen et al., 2024). However, the potential of ERQC to modulate EV-mediated protein trafficking is not yet systematically investigated.

This study presents the first systematic analysis of N-glycan-dependent ERQC in C. neoformans using mutant strains lacking ERQC gene homologs. Our findings highlight the critical roles of ERQC in maintaining cellular fitness and facilitating EV-mediated transport of virulence factors.

Results

Evolutionary unique features of C. neoformans ERQC components

We performed BLAST analysis of the C. neoformans H99 genome to identify the ERQC pathway-associated homologous genes of C. neoformans, followed by comparison with the ERQC components in other eukaryotic organisms (Supplementary Fig. S1, A). Unlike most eukaryotes, several yeast species within the Ascomycota phylum, such as Saccharomyces cerevisiae and Candida albicans, do not possess a functional UGGT. In contrast, UGGT homologs were identified in most Basidiomycota fungal species, including C. neoformans (Supplementary Fig. S1, A and B). The C. neoformans UGGT (CNAG_03648), which has been named as Ugg1, consists of 1,582 amino acids (aa) and features a signal peptide (1–20 aa) along with four tandem-like thioredoxin-like (TRXL) domains: TRXL12 (33–320 aa), TRXL13 (28–416 aa), TRXL14 (432–618 aa), and TRXL15 (712–950 aa). Ugg1 also contains a glucosyltransferase (GT) 24 domain with a DXD motif (1,369–1,371 aa), and a KDEL-like ER retention signal (1579–1582 aa), which facilitates its retrieval from the Golgi apparatus (Fig. 1B, Supplementary Fig. S2, A).

Additionally, we identified two C. neoformans ORFs, Mns1 (CNAG_02081) and Mns101 (CNAG_03240), as homologs of the eukaryote α1,2-mannosidase I, which processes N-glycans before exporting them to the Golgi. The C. neoformans Mns1 and Mns101 show 42.4% and 28.8% amino acid identities to S. cerevisiae ER α1,2-mannosidase I, respectively, and share 30.9% identity between them. Furthermore, C. neoformans Mnl1 (CNAG_01987) and Mnl2 (CNAG_04498) were identified as putative components of ERQC in C. neoformans (Supplementary Fig. S1, A and C). C. neoformans Mnl1 is a homolog of the yeast α1,2-mannosidase-like protein Htm1, which processes N-glycans targeted for ERAD. In contrast, C. neoformans Mnl2 encodes a mannosidase that does not share significant similarity with mannosidases from other eukaryotes. Mns1, Mns101, and Mnl1 possess glucosyl hydrolase (GH) 47 domains, essential for mannosidase activity, whereas Mnl2 contains a GH92 domain, also associated with mannosidase activity (Fig. 1B). The Mns1 and Mnl1 families are characterized by conserved cysteine (Cys) and alanine (Ala) residues within their activity domains, as previously described (Jakob et al., 2001). These conserved residues are present in C. neoformans Mns1, Mns101, and Mnl1 but are absent in Mnl2 (Supplementary Fig. S2, B). Notably, Mns101 and Mnl2 are Basidiomycota-specific proteins (Supplementary Fig. S1, A) and appear to have diverged early from other fungal mannosidase clades based on phylogenetic analysis (Supplementary Fig. S1, C). Interestingly, neither Mnl1 nor Mnl2 contains canonical KDEL/HDEL-like ER retention signals. In S. cerevisiae, the ER retention of Mnl1/Htm1 is mediated through its interaction with protein disulfide isomerase Pdi1, which carries an HDEL sequence (Gauss et al., 2011). Similarly, C. neoformans Mnl1 and Mnl2 may employ a non-canonical retention mechanism, likely facilitated by interactions with other ER-resident proteins, to achieve ER localization.

Loss of UGG1, MNS1, and MNS101 causes alteration of N-glycan profiles in C. neoformans

We performed high-performance liquid chromatography (HPLC) analysis of the cell wall mannoproteins (cwMPs) from both the wild-type (WT) and ERQC mutant strains to investigate the ERQC malfunction-induced structural differences in N-glycans (Fig. 2). The HPLC profiles of cwMPs from the WT strain showed an M8 peak as the major species. This peak corresponded to N-glycans with 8 mannose residues (Fig. 2A, top). The glycan structure at the M8 peak (Man8GlcNAc2) in the WT primarily corresponded to the Man7GlcNAc2 core N-glycan with an additional mannose residue that is linked via an α1,6-linkage and added in the Golgi apparatus (Park et al., 2012). In the ugg1Δ mutant, M8 was also the main N-glycan species, but the pools of hypermannosylated N-glycans (larger than the M11 peak) were markedly reduced (Fig. 2A, middle). The altered N-glycan profile of the ugg1Δ mutant was restored to that of the WT after complementation with the WT UGG1 gene (Fig. 2A, bottom). The lectin blotting analysis using Galanthus nivalis agglutinin (GNA), which specifically binds to terminal α1,2-, α1,3-, and α1,6-linked mannose residues, showed a distinctive increase of glycoproteins with lower molecular weight in the secretory proteins in ugg1Δ compared to those in WT (Fig. 2B). This observation aligns with the decrease of hypermannosylated N-glycans in the ugg1Δ mutant. Overall, these results suggest that Ugg1 is involved in mediating the hypermannosylation of N-glycans in the Golgi apparatus.

N-glycan profile analysis of C. neoformans ERQC mutant strains.

The HPLC and MALDI-TOF-based N-glycan structure analysis were carried out as described in Supplementary information. (A) HPLC-based analysis of N-glycan profiles of the ugg1Δ mutant. (B) Lectin blotting of sodium dodecyl sulphate (SDS)-polyacrylamide gels containing intracellular or secreted proteins into the culture supernatants of the wild type (WT) and ugg1Δ strains. Yeast cells were cultivated in YPD medium for 24 h, harvested, and subjected to sample preparation of soluble intracellular proteins and secreted proteins. The proteins (30 μg) were loaded on 15% SDS-polyacrylamide gel and analyzed using silver staining (left) or blotting (right) with Galanthus nivalis agglutinin conjugated to horseradish peroxidase (GNA-HRP, Roche). (C) HPLC analysis of total N-glycan profiles of mns1Δ, mns101Δ, and mns1Δ101Δ mutants. (D) MALDI-TOF profiles of neutral N-glycans of mns1Δ, mns101Δ, and mns1Δ101Δ mutants. The N-glycans of cell wall mannoproteins from C. neoformans cells were AA-labelled and analyzed using HPLC. For MALDI-TOF analysis, neutral N-glycan fractions were obtained from the HPLC fractionation of total N-glycans.

We also assessed putative α1,2-mannosidase I genes for roles in protein mannosylation. The mns1Δ mutant N-glycan profile showed a peak shift from an M8 to an M9 form, which strongly indicates that C. neoformans Mns1 could be the primary ER α1,2-mannosidase I (Fig. 2C, mns1Δ). Notably, the loss of Mns101, which is present only in Basidiomycota, increased the fractions containing hypermannosylated glycans (> M10) while maintaining M8 as a primary core N-glycan form. This suggests that the basidiomycete-specific Mns101 may potentially be a novel α1,2-mannosidase that functions to remove mannose residues from hypermannosylated N-glycans in the Golgi apparatus or further trims the M8 glycan in the ER before the glycoproteins are transported to the Golgi (Fig. 2C, mns101Δ). The matrix-assisted laser desorption ionization-time of flight (MALDI-TOF) mass spectrometer analysis of neutral N-glycans (Fig. 2C), further confirmed the shift of the major M8 peak in the WT strain to M9 in the mns1Δ mutant and the increase in the hypermannosylation profile in the mns101Δ mutant (Fig. 2D). The N-glycan profiles of the mns1Δ101Δ double mutant showed the combined effect of each null mutation, and it showed both increased hypermannosylated glycans and a shift from the M8 to the M9 peak in both the HPLC (Fig. 2C, mns1Δ101Δ) and MALDI-TOF analyses. This indicates that Mns1 and Mns101 serve as mannosidases and play independent roles at different stages of N-glycan processing in C. neoformans.

In contrast, the loss of both MNL1 and MNL2 did not show notable differences in the HPLC profile of N-glycans from cell surface mannoproteins (Supplementary Fig. S3, A). Considering the expected function of Mnl1 and Mnl2 in ERAD, their substrates are likely misfolded proteins that have not been transported to the Golgi apparatus. Thus, we speculate that only the normally processed N-glycan profiles of cell surface mannoproteins in the mnl1Δ mnl2Δ mutant strain were observed.

Loss of ERQC components results in defective growth fitness and increased stress sensitivity

To investigate the changes in ERQC-related gene expression under stress conditions, we conducted quantitative reverse transcription PCR (qRT-PCR) analysis of C. neoformans following treatment with tunicamycin (TM, 5 µg/ml), a well-established inhibitor of N-glycosylation, or dithiothreitol (DTT, 20 mM), which disrupts disulfide bond formation and induces the accumulation of misfolded proteins. Additionally, we examined gene expression after growth at 37 °C (Fig. 3A). When misfolded proteins accumulate in the ER, the unfolded protein response (UPR) system is activated to restore homeostasis (Chakraborty et al., 2016). As expected, KAR2, encoding a molecular chaperone associated with the UPR system, was upregulated in response to TM treatment, DTT treatment, and heat stress at 37 °C. The ERQC components UGG1, MNS1, MNS101, MNL1, and MNL2 were upregulated by DTT treatment; however, only UGG1 was slightly induced following TM treatment. The high-temperature conditions also induced the expression of KAR2, UGG1 and MNL2. Notably, DTT treatment induced a 10-fold higher expression of MNS1 and MNL1. These results strongly suggest that the C. neoformans genes UGG1, MNS1, MNS101, MNL1, and MNL2 are crucial components of the ERQC and ERAD systems, although each respond uniquely to various cell/ER stress conditions.

Growth phenotype of C. neoformans ERQC mutant strains.

(A) Expression analysis of ERQC genes in C. neoformans. Yeast cells were cultured in YPD medium to a mid-logarithmic phase and exposed to dithiothreitol (DTT; 20 mM), tunicamycin (TM; 5 µg/ml) or cultured at 37 °C for 1 h. The relative transcript levels of C. neoformans genes were analyzed using qRT-PCR and normalized with that of ACT1. Error bars represent standard deviation of duplicated assays. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. *** P < 0.0005, ** P < 0.003, *** P < 0.005, * P < 0.05. (B) Spotting analysis of C. neoformans ugg1Δ mutant strains under various stress conditions such as heat stress (37 °C and 39 °C), ER stress (DTT and TM), cell-wall stress (CFW: calcofluor white, CR: Congo red, SDS: Sodium dodecyl sulfate, caffeine), osmotic stress (NaCl, KCl, sorbitol) and treatment with antifungal drugs (fluconazole, ketoconazole, fludioxonil). (C) Growth analysis in the presence of 5’,5’,5’-trifluoroleucine (TFL). Respective strains were spotted on SC-Leucine media with or without TFL supplementation. Plates were incubated for 3 days at 30 °C. (D) RT-PCR analysis of IRE1-dependent splicing of HXL1. Strains were cultured in YPD supplemented with 5 µg/ml TM.

We next investigated the growth of the ugg1Δ mutant under various stress-inducing conditions to elucidate the roles of the ERQC components in ER stress response and adaptation, which are closely associated with C. neoformans virulence. The ugg1Δ mutant showed impaired growth even under normal growth conditions (Fig. 3B) and increased sensitivity to ER stress-inducing agents such as DTT, TM and cell wall stressors, including as calcofluor white (CFW), congo red (CR), sodium dodecyl sulfate (SDS), and caffeine. Additionally, ugg1Δ showed hindered growth in the presence of antifungal drugs such as azole agents (fluconazole and ketoconazole) and the glucose transport inhibitor fludioxonil. These findings suggest that Ugg1 is crucial for the robust growth and survival of C. neoformans under various stress conditions. In contrast, the single deletion (mns1Δ and mns101Δ) and double deletion (mns1Δ101Δ) strains did not exhibit noticeable phenotypic changes except for a slight increase in sensitivity of mns101Δ and mns1Δ101Δ to higher fludioxonil concentrations (Supplementary Fig. S3, B). Similarly, neither the single nor double disruption of MNL1 and MNL2 produced detectable changes in the tested growth conditions (Supplementary Fig. S3, C).

Amino acid analogues have been used for evaluating the functionality of protein folding, as its incorporation into newly synthesized proteins interrupts proper protein folding (Trotter et al., 2002; Cowie et al., 1959). We hypothesized that, in the context of a defective ERQC pathway, misfolded proteins cannot be adequately repaired and accumulate, thus triggering ER stress, which may ultimately inhibit cell growth in the presence of amino acid analogues. In the presence of the leucine analogue 5’,5’,5’-trifluoroleucine (TFL), ugg1Δ showed noticeably inhibited growth, whereas the WT strain showed no growth inhibition (Fig. 3C). The mns1Δ101Δ strain also exhibited slightly increased sensitivity to TFL compared with that of the WT strain.

As a defense mechanism against misfolded protein accumulation-mediated ER stress, UPR is induced by the unconventional splicing of the HXL1 transcription factor in C. neoformans (Cheon et al., 2011). RT-PCR analysis revealed a significantly elevated level of spliced HXL1 in the ugg1Δ strain compared to the WT, even under normal culture conditions (Fig. 3D), strongly indicating that the loss of UGG1 leads to the accumulation of misfolded proteins, triggering ER stress under standard growth conditions. Furthermore, green fluorescence protein (GFP)-tagged Ugg1, Mns1, and Mns101 proteins colocalized with the ER marker, supporting their role as confirming them to be functional ERQC components based on their subcellular localization in the ER (Supplementary Fig. S4).

ERQC defects lead to virulence attenuation in C. neoformans

We investigated the effects of a defective ERQC on in vitro virulence phenotypes of C. neoformans by analyzing capsule and melanin production, as these are two of the major virulence factors of C. neoformans. Culturing on L-DOPA-containing plates showed that melanin production was reduced in the ugg1Δ strain compared with that in the WT (Fig. 4A, top), whereas the mns1Δ101Δ mutant cells did not show detectable defects at either 30 or 37 °C (Fig. 4A, bottom). Considering the decreased growth of the ugg1Δ strain, the melanin production of C. neoformans cells cultivated in liquid L-DOPA was measured and normalized by cell density, which further confirmed the decreased melanin production activity of the ugg1Δ strain (Fig. 4B). India ink staining of the capsule showed significantly reduced capsule thickness in the ugg1Δ strain and a moderate defect in the mns1Δ101Δ strain (Fig. 4C).

In vitro and in vivo virulence-associated phenotypes of C. neoformans UGG1, MNS1, and MNS101 mutant strains.

(A) Melanin synthesis analysis on L-DOPA plates. WT, ugg1Δ, ugg1Δ::UGG1, mns1Δ, mns101Δ, mns1Δ101Δ, mns1Δ101Δ::MNS1, mns1Δ101Δ::MNS101, and cac1Δ (negative control) strains were serially diluted, plated on L-DOPA plates, and incubated at 30 °C and 37 °C. (B) Melanin synthesis activity per cell of the ugg1Δ mutant. WT, ugg1Δ, ugg1Δ::UGG1 were cultured in liquid L-DOPA medium. The amount of melanin in the culture supernatant was measured and normalized by cell density. Capsule formation. Cells were cultured for 2 days in 10% Sabouraud media at 30 °C and observed under the microscope. Statistical significance: ****, P < 0.0001, ns, not significant). In vivo virulence analysis. A/Jcr mice (n=8) were infected with 105 cells of WT, ugg1Δ, and ugg1Δ::UGG1, mns1Δ, mns101Δ, mns1Δ101Δ, mns1Δ101Δ::MNS1, and mns1Δ101Δ::MNS101 strains, and survival was monitored for 2 months, as described in the Supplementary Information. (E) Survival of C. neoformans in macrophages. Survival of C. neoformans cells within the J774A.1 macrophage-like cell line was determined by counting colony formation unit (CFU) obtained from lysed macrophages from two biologically independent experiment sets. **** P < 0.0001, *** P < 0.0005, * P < 0.05, ns, not significant. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test.

Defects in the ERQC system led to a more apparent decrease in pathogenicity in a murine model of systemic cryptococcosis. Notably, the ugg1Δ strain was almost avirulent with no detectable signs of illness in the infected animals (Fig 4D). Although MNS1 and MNS101 single disruptions did not cause a detectable decrease in C. neoformans pathogenicity, their double disruption resulted in decreased virulence. Histopathological analysis of infected mice lungs suggested significantly reduced lung colonization by both ugg1Δ and mns1Δ101Δ cells (Supplementary Fig. S5, A). Additionally, the fungal burden of ugg1Δ-infected animals showed a notable decrease in organ colonization at 60 days post-infection (dpi) (Supplementary Fig. S5, B). Analysis of the fungal burden at 7 dpi showed significantly reduced organ colonization of the mns1Δ101Δ mutant compared with that of the WT (Supplementary Fig. S5, C). Consistent with the in vivo survival data, the number of surviving ugg1Δ and mns1Δ101Δ cells within the macrophage-like cell line J447A.1 was significantly lower compared with that of the WT strain (Fig. 4E). These data suggest that a functional ERQC system is critical for maintaining full in vivo pathogenicity and ensuring robust survival within host immune cells.

Loss of Ugg1 function manifests defects in extracellular transport of GXM

Cryptococcal capsule materials are synthesized intracellularly and secreted to the cell surface. Subsequently, they are assembled and bound to the cell wall (Yoneda and Doering, 2006). The observed capsule defects in the ugg1Δ and mns1Δ101Δ mutant strains may be attributed to issues in the capsule polysaccharide synthesis or transport steps. Alternatively, defective capsule formation could arise from an increased shedding of capsule polymers, preventing their proper attachment to the cell surface. To determine the cause of the capsule defects, we examined the amounts of capsular polysaccharides synthesized intracellularly and those shed into the culture supernatant by a capsule blotting assay, using the 18B7 antibody directed against GXM, the major capsule component (Casadevall et al., 1998). As controls, we also tested capsule production and secretion in the cap59Δ strain with a defect in GXM synthesis (Grijpstra et al., 2009) and the rim101Δ strain with a capsule attachment defect despite normal polysaccharide synthesis and secretion (O’Meara et al., 2010). Intracellular GXM quantities did not notably differ between the WT, ugg1Δ, mns1Δ101Δ, and rim101Δ strains, although ugg1Δ showed slightly reduced GXM level (Fig. 5A, left). In contrast, no intracellular GXM was detected in cap59Δ, which was consistent with its defect in GXM synthesis. Moreover, the ugg1Δ strain was defective in capsule polymer secretion/shedding, whereas the mns1Δ101Δ strain displayed enhanced shedding of GXM-containing polymers that appeared to be shorter in length than the WT-type polymers, suggestive of poor polysaccharide polymerization (Fig. 5A, right). Therefore, these results suggest that ugg1Δ contains a defect in polysaccharide trafficking to the extracellular space, whereas mns1Δ101Δ has a defect in polysaccharide polymerization, resulting in defective capsule elaboration, albeit with differential degrees of severity in both ERQC mutants.

Capsule shedding and transfer analysis of C. neoformans UGG1, MNS1, and MNS101 mutant strains.

(A) Capsule shedding analysis. The presence of intracellular (left) and shed (right) glucuronoxylomannan (GXM) was assessed by blotting a cell culture filtrate using the monoclonal antibody 18B7. The arrow indicates the direction of electrophoresis. (B) Capsule transfer analysis using exogenous capsule material from WT. The capsule transfer assay was performed using the indicated strains as acceptors. Surface capsules were probed using the anti-GXM antibody 18B7 conjugated with AlexaFluor 488. Quantitative measurement of fluorescence intensity was calculated based on independent triplicate experiments with standard deviations presented as error bars. (C) Capsule transfer analysis using exogenous capsule material from WT or mns1Δ101Δ. Statistical significance: **** P < 0.0001. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. (D) Transmission electron microscopy (TEM) of C. neoformans WT, ugg1Δ, and mns1Δ101Δ strains. Yeast cells were grown overnight at 30 °C in YPD medium and fixed in 2% glutaraldehyde and 2% paraformaldehyde. A Zeiss Axioscope (A1) equipped with an AxioCan MRm digital camera was used to visualize India ink-stained C. neoformans cells. Specimens were prepared using critical point drying prior to TEM microscopy. Capsule and yeast cell body diameters were measured using ImageJ (National Institute of Health).

Furthermore, to determine whether the ERQC mutants are also impaired in capsule attachment, we performed a capsule transfer assay, to assess whether exogenously shed capsule polysaccharides can bind to acapsular mutants (Reese and Doering, 2003). The hypocapsular rim101Δ mutant did not show reassociation of capsular polysaccharides; however, the acapsular cap59Δ mutant showed recovery of capsule formation when co-incubated with exogenous GXM (Fig. 5B). Similarly, ugg1Δ reverted its acapsular phenotype on incubation with the WT-shed polysaccharides, strongly suggesting that the defective capsule phenotype of ugg1Δ is primarily due to defective GXM secretion rather than impaired attachment of capsule polymers. We further conducted a capsule transfer assay using capsular polysaccharides shed from mns1Δ101Δ. The cap59Δ mutant reverted to the capsular phenotype by attaching to the mns1Δ101Δ-shed polymers. However, the fluorescence intensity of the capsule generated was significantly lower than that of the cells attached with polysaccharides obtained from the WT. Although the decreased length of capsule polysaccharides should be validated by techniques specifically measuring GXM size (De Jesus et al., 2010), this result suggests that the mns1Δ101Δ double mutant secretes incomplete capsule polysaccharides (Fig. 5C), leading to a hypocapsular phenotype. These results collectively indicate that disrupting the ERQC results in impaired capsule formation by defective GXM trafficking to the extracellular space.

Transmission electron microscopy (TEM) imaging of the WT and mutant cells distinctly showed a diminished capsule structure and loss of capsule shedding in the ugg1Δ strain compared with that of the WT strain (Fig. 5D). Additionally, we observed considerable thinning of the cell walls in both ERQC mutant cells. Notable changes in intracellular structures along with an increase in the number of pigmented vesicles were observed in the mutant cells even under YPD culture conditions. Under capsule-inducing conditions, a significant accumulation of vesicular structures (electron-lucent structures) was observed intracellularly, particularly in ugg1Δ. We performed lipid droplet (LD) staining to determine whether these vesicular structures might be LDs. We observed a notable increase in LDs in ugg1Δ under capsule-inducing culture conditions (Supplementary Fig. S6, A). LDs impact proteostasis by sequestering misfolded proteins intended for degradation and providing an “escape hatch” when the ERQC is overloaded (Ploegh, 2007; Vevea et al., 2015). Additionally, FM4-64 dye staining showed increased number of vacuoles in the ugg1Δ cells compared with that in the WT (Supplementary Fig. S6, B). Moreover, all the observed abnormal phenotypes were less pronounced in the mns1Δ101Δ mutant than in the ugg1Δ strain. Collectively, the altered vesicular structures observed in the ugg1Δ and mns1Δ101Δ strains highlight the critical role of ERQC in maintaining proper vesicular architecture and cell surface organization. These functions are tightly linked to extracellular trafficking of virulence factors and cell wall remodeling.

ugg1Δ transcriptomic profiling shows induced ER and cell wall integrity stress responses

To elucidate the mechanisms underlying UGG1 deletion-induced physiological changes in C. neoformans, we performed RNA sequencing (RNAseq)-based transcriptome analysis of the WT and ugg1Δ strains under standard growth conditions (YPD medium, 30 °C). Comparative transcriptome analysis showed statistically significant alterations exceeding 2-fold in the expression patterns of 146 genes, of which 85 were upregulated and 61 were downregulated compared with that of the WT strain (Fig. 6A, B, Supplementary Tables S2, A and B).

Transcriptome analysis of C. neoformans WT and ugg1Δ cells.

(A) Volcano plot comparing a 2-fold differential gene expression between ugg1Δ and WT strains under normal growth conditions. (B) Number of genes upregulated and downregulated by ≥ 2-fold in ugg1Δ compared with that of the WT. (C) Gene ontology (GO) analysis of differentially expressed genes between WT and ugg1Δ strains. Significantly upregulated genes in ugg1Δ are shown in red, whereas significantly downregulated genes in ugg1Δ are shown in blue. Total RNA was extracted and subjected to RNAseq analysis as described in the Supplementary Information. (D) qRT-PCR analysis of mRNA expression levels of a set of genes responsible for capsule biosynthesis, cell wall remodeling, and both conventional and non-conventional secretion in ugg1Δ vs WT under normal growth conditions from three biologically independent experiment sets.

Notably, the ugg1Δ mutant showed upregulation of genes such as SKN1 and KRE6 (transmembrane glucosidases involved in the sphingolipid biosynthesis and β-glucan biosynthesis), CAT2 (a putative peroxisomal catalase), CHS7 (chitin synthase export chaperone), and CNAG_05458 (a putative endo-1,3(4)-β-glucanase). In contrast, genes encoding ERG3 and ERG25 (associated with the ergosterol biosynthesis pathway) and FHB1 (flavohemoglobin denitrosylase associated with counteracting nitric oxide stress) were significantly downregulated. Gene Ontology (GO) analysis in ugg1Δ showed pronounced induction of genes implicated in various cellular processes, including the hydrolysis of O-glycosyl compounds for cell wall remodeling, DNA replication, proteolysis, ribosome biogenesis, protein folding, and serine/threonine kinase activity. In contrast, genes associated with carbohydrate and lipid metabolism, chaperone binding, iron homeostasis, and mitochondrial intermembrane space were considerably downregulated (Fig. 6C). The transcriptome profile strongly suggests that loss of Ugg1 function induces the expression of several genes involved in maintaining cell wall integrity to compensate for cell wall defects, particularly those associated with chitin (CDA2, CHS7, QRI1, and CNAG_06898) and glucan biosynthesis (SKN1, KRE6, EBG1, LPI9, CNAG_05458, and BLG2). Furthermore, genes coding for ER chaperones that aid in protein folding (ERO1, KAR2, LHS1, and PDI1) and chaperone regulator (SCJ1), and genes involved in proteolysis in the ER (CNAG_04635, CNAG_06658) were induced, likely as part of the ER stress response induced by the accumulation of misfolded proteins in the presence of a defective ERQC.

Further investigation of the effects of UGG1 disruption in the expression of genes involved in capsule biosynthesis, cell wall remodeling, and the secretion pathways of protein virulence factors using qRT-PCR analysis (Fig. 6D) showed no noticeable changes in the expression of genes related to capsule biosynthesis and conventional or unconventional secretion pathways at mRNA level, despite the aberrantly displayed defective phenotypes.

EV-mediated protein trafficking and non-conventional secretion is impaired in the ugg1Δ mutant strain

Many enzymes contribute to the composite cryptococcal virulence phenotype. Some of these virulence-associated enzymes are secreted through traditional secretion pathways, whereas others are packaged into EVs and released into the extracellular milieu via non-conventional secretion mechanisms (Almeida et al., 2015). The defects observed in the ERQC mutant strains, particularly in the production of melanin and polysaccharide capsules, both of which are transported by EVs, suggest a possible impairment in the EV-mediated extracellular transport of virulence-associated enzymes in these mutants. Therefore, we examined the in vitro activity of urease, and laccase and acid phosphatase in both intracellular and extracellular fractions.

Comparative analysis of the WT and ugg1Δ strains showed minimal changes in the intracellular activities of urease, laccase, and acid phosphatase. However, their activities were significantly reduced in the extracellular fractions of the ugg1Δ cells (Fig. 7A, 7B, 7C, and 7D). Importantly, urease activity was performed in both solid and liquid media, demonstrating a notable loss of secreted urease activity in ugg1Δ cells, with a slight decrease in the mns1Δ101Δ double mutant cells (Fig. 7A and 7B). As urease lacks a signal peptide required for conventional protein secretion, reduced urease secretion strongly indicates a compromised EV-mediated non-conventional secretory pathway in the ugg1Δ mutant strain. The decrease in extracellular laccase activity (Fig. 7C) corroborates our melanization assay results, in which ugg1Δ showed defective melanin production (Fig. 4A).

Analysis of protein secretion in C. neoformans UGG1, MNS1, and MNS101 mutant strains.

(A) Spot assay for urease analysis on Christensen’s urea agar. Absence of pink coloration indicates loss of urease activity. (B, C, D) Analysis of secretion for virulence-related enzymes such as urease, laccase, and acid phosphatase from three biologically independent experiment sets. Statistical significance: **** P < 0.0001, ** P < 0.003, *** P < 0.005, * P < 0.05. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. (E, F) Analysis of secretion for non-virulence-related enzymes such as cellulase and α-amylase. Statistical significance: *** P<0.0005, ** P<0.003 * P < 0.05, ns, not significant. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. (G) Analysis of the conventional secretion of Cda1 in C. neoformans. Presence of Cda1 was analyzed in total (T), soluble (S), and insoluble (I) fractions of intracellular extracts (left), along with the secreted fraction (right). Subcellular fractionations were performed as previously described (Thak et al., 2022), and the fractions were subjected to western blotting analysis using an anti-Cda1 antibody.

We next examined cellulase and α-amylase activities to evaluate the impact of ERQC disruption on the conventional secretion pathway of non-virulence-related enzymes (Fig. 7E and 7F). The extracellular activities of these carbohydrate polymer-degrading enzymes in the ugg1Δ mutant were slightly lower than those observed in the WT, likely reflecting reduced intracellular enzyme production rather than a substantial impairment of secretion efficiency. Additionally, we analyzed the localization of chitin deacetylase I (Cda1), a glycosylphosphatidylinositol (GPI)-anchored protein involved in converting chitin to chitosan, essential for maintaining cell wall integrity (Baker et al., 2007; Baker et al., 2011; Upadhya et al., 2018; Upadhya et al., 2021). Cda1 was primarily detected in the insoluble cellular protein fraction, encompassing the cell wall and membrane compartments (Fig. 7G, left), consistent with its cell surface localization via a GPI anchor. Cda1 is secreted extracellularly upon cleavage of the GPI anchor. Notably, the ugg1Δ mutant exhibited significantly higher intracellular and extracellular Cda1 levels compared to the WT and mns1Δ101Δ strains (Fig. 7G, right), aligning with a 1.87-fold increase in CDA1 mRNA expression observed in the RNA-seq data. These findings indicate that the surface localization and secretion of Cda1 remain efficient, suggesting that the conventional secretion pathway is largely unperturbed in ugg1Δ. Collectively, these results suggest that while Ugg1-mediated ERQC defects have a pronounced negative effect on EV-mediated protein transport, their impact on the conventional secretion pathway is minimal.

Extracellular vesicle biogenesis and cargo loading are defective in ugg1Δ strain

We thus analyzed the number and size distribution of EVs to determine possible abnormal EV-mediated trafficking of virulence factors in the ugg1Δ mutant. Nanoparticle tracking analysis (NTA) showed a major peak in size distribution at approximately 150 (134±28) nm in the WT strain, which was similar in size to mammalian exosomes and typical microbial EVs. Additionally, a minor peak ranging from 300–500 nm was observed and corresponded to microvesicles (Fig. 8A). Notably, the size distribution of ugg1Δ EVs was more heterogeneous with smaller EVs ranging from 50–150 nm (82±16; 124±14) compared with that of the WT. The cap59Δ EVs displayed a major distribution of approximately 150 nm (120±21). Microvesicles were barely detected in either cap59Δ or ugg1Δ strains.

Analysis of extracellular vesicles purified from WT, ugg1Δ, and cap59Δ cells.

(A, B) Nanoparticle tracking analysis (NTA) of EVs extracted from WT, ugg1Δ, and cap59Δ strains and quantification of total extracellular vesicle (EV) concentration per cell density. Quantitative measurements were derived from three independent experiments with standard deviations presented as error bars. Statistical significance: *** P<0.0005, * P< 0.05. All statistical data were determined based on one-way ANOVA and Dunnett’s post-hoc test. (C, D) Cryo-TEM imaging of purified EVs and comparative analysis of EV size in WT, ugg1Δ, and cap59Δ strains. Scale bar, 100 nm. The outer EV diameter of a total number of 100 EVs per strain, captured using cryo-TEM, were measured. (E) Heatmap representation of fold change between WT and ugg1Δ EV-associated proteins, commonly detected in this study and in previously reported EV proteome datasets (ugg1Δ/WT). Upregulated proteins in ugg1Δ are shown in red, whereas downregulated proteins are shown in blue. The proteome data of whole-cell lysates (WCL), generated from the cell pellets obtained after EV separation, were included for comparison. (F) Blotting analysis of GXM in the C. neoformans cells and EVs of WT, ugg1Δ, and cap59Δ strains. The 8 M urea extracts were obtained from EVs and cell pellets, from which EVs are generated. The urea extracts (5 μg total proteins) were loaded on 8% SDS-polyacrylamide gel and subjected to silver staining or blotting analysis using the anti-GXM 18B7 (α-GXM) and anti-Cda1 (α-Cda1) antibodies, respectively. Left: total cell extract. Right: total EV extract.

We observed a significant reduction in the total number of secreted EVs in the ugg1Δ mutant (approximately 40%), suggesting defective EV biogenesis and/or stability in the absence of functional Ugg1 (Fig. 8B). Cryo-TEM analysis of EV morphology confirmed that the sizes of EVs released by ugg1Δ were significantly smaller and more diverse than those of the WT or the acapsular mutant (Fig. 8C), further supporting the NTA results. Examining EV size distribution by measuring their diameter (Fig. 8D) further confirmed that although the EVs in the WT strain ranged from approximately 50–550 nm in size with distribution primarily concentrated at approximately 150 nm, the EVs in the ugg1Δ mutant were smaller (< 100 nm) with heterogeneous sizes. Notably, cap59Δ EVs were uniformly distributed at approximately 150 nm and were present at a higher concentration than those of EVs from the WT. An increase in EV release and virulence factors has been also observed in acapsular mutant strains of C. neoformans and C. gattii (Rodrigues et al., 2007; Reis et al., 2019), supporting the notion that the capsule serves as a barrier in EV release in capsule forming species.

To characterize the protein composition of EVs from WT and ugg1Δ strains, EV-associated and whole-cell lysate (WCL) proteins were extracted and subjected to proteomic analysis (Fig. 8E). In the WCL, 4,678 proteins were identified in the WT strain, with 333 exhibiting differential expression (>2-fold) in ugg1Δ versus the WT strain, including 123 upregulated and 210 downregulated proteins (Supplementary Fig. S7, A, left). In EVs, 2,075 proteins were detected in the WT strain, of which 693 were differentially expressed in EVs from ugg1Δ relative to EVs from the WT strain (>2-fold), comprising 273 upregulated and 420 downregulated proteins (Supplementary Fig. S7, A, right; Supplementary Table S3). Additionally, comparative proteomic analysis revealed 180 proteins uniquely enriched in EVs (Supplementary Fig. S7, B). The GO enrichment analysis of differentially expressed proteins (> 2-fold) showed distinctive patterns in biological process (BP) and cellular components (CC) between EVs and the WCL (Supplementary Fig. S7, C). Three proteomic analyses of C. neoformans EVs have been reported previously (Rodrigues et al., 2008; Wolf et al., 2014; Rizzo et al., 2021), identifying 76, 202, and 1,847 proteins associated with EVs, respectively. We compared the EV protein content between WT and ugg1Δ identified in our study, focusing on overlapped proteins from the three EV proteomic data sets reported. Among 88 commonly reported EV-associated proteins, 29 were downregulated while 16 were overexpressed in ugg1Δ EVs compared to WT EVs (Fig. 8E; Supplementary Table S4). The differential expression pattern of the EV-associated proteins is quite distinctive from that of WCL, revealing significantly detectable changes in EVs. Thus, the absence of functional Ugg1 not only caused morphological alterations and reduced EV numbers but also altered EV protein abundance in C. neoformans.

To further examine possible cargo loading defects, we investigated the presence of GXM in the ugg1Δ EVs, comparing GXM quantities in the total cell extracts from WT, ugg1Δ, and cap59Δ strains (Fig. 8F). Cda1 was used as the representative EV-associated protein as its presence was reported within the C. neoformans EV membrane, despite a substantial portion of Cda1 being secreted through the conventional secretion pathway (Rizzo et al., 2021). GXM polysaccharides were not detected in the EVs released from ugg1Δ, although it was present in the whole-cell extract. Taken together with the reduced quantity of EV-associated proteins, the absence of GXM polysaccharides in ugg1Δ EVs strongly indicates that EV cargo loading is defective when the ERQC is dysfunctional in C. neoformans.

Discussion

As glycoproteins pass through the ER-Golgi secretory pathway, their N-glycans undergo extensive modifications in association with protein quality control. The ERQC system is highly conserved among eukaryotes but also diverged with distinct and species-specific features. The ERQC composition of C. neoformans is unique in that it possesses UGGT but lacks the three glucosyltransferases necessary for the addition of glucose residues to the core precursor N-glycans (Park et al., 2012). Additionally, it lacks CRT, and carries multiple α1,2-mannosidases. In this study, we investigated the molecular features and functions of the N-glycan-dependent ERQC in C. neoformans, which generates unique N-glycan precursors without glucose addition, and shorter in length than those of most eukaryotes. Our data strongly suggest that despite the incomplete composition of the ERQC components, the UGGT-centered ERQC plays pivotal roles in cellular fitness, and particularly in the EV-mediated extracellular transport, which is crucial for pathogenicity.

The virulence of the C. neoformans ugg1Δ mutant was almost abolished in mice (Fig. 4D), a phenotype consistent with the observed defects in key virulence determinants, such as the capsule and melanin, as well as poor growth at 37 °C (Fig. 3B; Fig. 4A and 4B). The mns1Δ101Δ mutant displayed intermediate phenotypes between those of the WT and ugg1Δ. Our data from the UPR induction analysis and comparative transcriptomic analysis strongly indicate that the ERQC mutation generates ER stress; this accounts for significant similarity of the defective phenotypes of the C. neoformans ERQC mutants to those reported in other mutants that exhibit defective protein folding in the ER, particularly in terms of increased stress sensitivity and decreased virulence. Connections between ER stress and thermotolerance have previously been established in C. neoformans, as growth at 37 °C requires key ER protein chaperones and protein processing machinery, including components of the UPR signaling pathway Ire1 and the ER stress-responsive transcription factor Hxl1 (Cheon et al., 2011; Havel et al., 2011; Jung et al., 2013). Other mutations causing defects in ER function in C. neoformans, such as mutants lacking DNJ1 (an ER J-domain containing co-chaperone) and CNE1 (an ER chaperone), resulted in growth inhibition. The dnj1Δ mutant strain displayed impaired elaboration of virulence factors, such as the exopolysaccharide capsule and extracellular urease activity, when cultured at human body temperature (Horianopoulos et al., 2021). Altogether with our data from the ERQC mutants, these findings strongly support the notion that maintaining ER homeostasis is crucial for survival and virulence factor production at elevated temperatures.

It is quite notable that the C. neoformans ugg1Δ EVs exhibited defective loading of GXM and many protein cargo, alongside a decrease in the number and changes in the size distribution (Fig. 8). Combined with compromised cell fitness, these defects in EV biogenesis and cargo loading, even under normal growth conditions, likely contribute to the complete loss of virulence owing to the defective export of virulence factors in the ugg1Δ mutant. Interestingly, our comparative proteome analysis of culture supernatants (secretome) revealed that the ugg1Δ strain exhibited more pronounced defects in the secretion of proteins lacking signal peptides, which are often linked to unconventional secretion mechanisms, compared to those with canonical secretion signals (Supplementary Fig. S8; Supplementary Table S5). These findings further support the role of ERQC in modulating non-conventional secretion pathways.

It is intriguing how the absence of Ugg1, leading to ERQC defects, results in defective EV biogenesis and cargo loading in C. neoformans. In fungi, EV production and release occur through the maturation of endosomes into multivesicular bodies (MVBs), which are directed to the cell surface. Upon fusion with the plasma membrane, MVBs release vesicles into the extracellular environment, functioning as an unconventional secretion pathway (Oliveira et al., 2010). MVBs formation relies on the functionality of the endosomal sorting complex required for transport (ESCRT), a highly intricate pathway involving a series of finely regulated events (Henne et al., 2011). Deletion of several ESCRT complex-related genes involved in EV transport directly impacts the cryptococcal capsule, notably resulting in reduced capsule size. Specifically, the C. neoformans vps27Δ strain showed altered EV size distribution, reduced capsule dimensions, defects in laccase export to the cell wall, and poor extracellular export of urease (Park et al., 2020). Other regulators of unconventional secretion are also linked to EV biogenesis. For example, the Golgi reassembly and stacking proteins (GRASPs) regulate EV cargo and dimensions in C. neoformans (Peres de Silva et al., 2018). A graspΔ mutant strain produced EVs with dimensions that significantly differed from those produced by WT cells, along with attenuated virulence and abnormal RNA composition. Moreover, the GRASP protein Grasp homology 1 (Grh1) serves as a chaperone that directly influences EV cargo (Malhotra, 2013; Peres da Silva et al., 2018). Autophagy regulators, which participate in EV formation in other eukaryotes, also play a role in cryptococcal EV formation. An atg7Δ strain manifests hypovirulence, and EVs produced by this strain show slightly different RNA composition compared with that of the WT cells (Oliveira et al., 2016). Notably, our data of the transcriptome and proteome analysis did not show significant changes in the ESCRT complex and GRASPs.

In addition to protein folding and secretion, the ER is crucial for regulating lipid metabolism (Moncan et al., 2021). Consequently, ER stress significantly impacts lipid and sterol synthesis, although some of these mechanisms are yet to be clarified. LDs aid the UPR and ERAD in degrading misfolded proteins during ER stress in S. cerevisiae (Garcia et al., 2021). In the present study, we observed a drastic increase in LDs in the ugg1Δ mutant (Supplementary Fig. S6, A), consistent with the previous hypothesis that LDs help maintaining ER homeostasis. Furthermore, we observed increased number of vacuoles in ugg1Δ (Supplementary Fig. S6, B). This may thus indicate abnormal lipid homeostasis caused by the ERQC defects, which could, in turn, affect EV biogenesis. The importance of lipids and membrane regulators in proper EV formation and GXM export has been suggested in a previous study on the Apt1 flippase in C. neoformans (Rizzo et al., 2018). Additionally, ergosterol is essential for membrane fluidity, permeability, and protein transport (Ermakova and Zuev, 2017). Erg6 is involved in the ergosterol biosynthesis pathway and was identified as essential for the trans-Golgi network transport of proteins (Proszynski et al., 2005; Nes et al., 2009). In C. neoformans, the erg6Δ mutant released EVs with a significantly larger diameter than those of the WT, carrying increased levels of proteins and sterols, highlighting the role of ergosterol in cryptococcal EV biogenesis (Oliveira et al., 2020). Notably, the ugg1Δ mutant exhibited loss of microvesicles, which are derived from the plasma membrane (Fig. 8A). Our preliminary analysis data on the surface lipid fraction of the ugg1Δ mutant indicated detectable alteration of sphingolipids and sterol profiles. Altogether, these findings suggest that ER stress caused by misfolded glycoprotein accumulation in ERQC-defective mutants may alter lipid composition, which could affect EV biogenesis.

A recent study using kidney cells presented strong evidence for a key role of ER stress in modulating EV biogenesis, demonstrating that ER stress decreases exosome production in mice (Fukuoka et al., 2023). They showed that T-cadherin is downregulated by ER stress through IRE1α activation at mRNA and protein levels and that ER stress decreases EV production through adiponectin/T-cadherin-independent way, which may involve interferon pathway activation in mice. We similarly observed induced activity of the ER stress sensor Ire1 in ugg1Δ, even under normal growth conditions, suggesting a possible association between the Ire1-mediated UPR pathway and EV biogenesis (Fig. 9). Additionally, glycosylation regulates the biogenesis of small EVs and affects protein cargo loading efficiency in melanoma cells (Harada et al., 2020). This suggests that the altered N- glycosylation observed in C. neoformans ugg1Δ may influence cargo loading of certain EV-targeting glycoproteins. Despite their findings not being conducted in fungi, they provide valuable guides on the investigation of conserved mechanisms shared with C. neoformans. Further studies to investigate the mechanisms underlying the ERQC-mediated modulation of EV biogenesis and cargo loading in C. neoformans will provide insights into understanding the regulation, production, composition, and diversity of fungal EVs, enabling a better understanding of their biological function. Expanding our knowledge on pathogenic fungal EVs would pave the way for utilizing native or engineered EVs as promising candidates for therapeutic applications, including fungal infection diagnosis and vaccine development.

Impact of ERQC disruption on glycoprotein folding and EV-mediated transport of virulence factors in C. neoformans.

(A) In WT strain, C. neoformans UGGT homolog, Ugg1, functions as a sensor for misfolded glycoproteins within the ER, playing a crucial role in protein quality control. Functional ERQC is essential not only for ensuring the proper folding of glycoproteins, which is critical for maintaining cellular fitness, but also for facilitating EV-mediated secretion of capsule polysaccharides and virulence-related enzymes necessary for pathogenicity. (B) In the UGGT-deficient strain (ugg1Δ), ER stress is increased because of misfolded protein accumulation within the ER lumen. This heightened stress leads to decreased cellular fitness, which negatively impacts EV biogenesis and cargo loading. Consequently, significant defects occur in EV-mediated transport, which ultimately leads to a complete loss of virulence. Nc: nucleus, Vc: vacuoles. This figure was partially created using BioRender (https://BioRender.com/3gzzput, https://BioRender.com/bgibgte).

Materials and Methods

Strains, culture conditions, plasmids, and primers

The C. neoformans strains that were constructed and used in this study are listed in Supplementary Table S1A. The plasmids and primers used are listed in Supplementary Tables S1B and S1C, respectively. The construction of the deletion mutants is described in the Supplementary Information. The strains were typically cultured in YPD medium (1% yeast extract, 2% bacto peptone, and 2% glucose) at 30 °C with shaking (220 rpm). C. neoformans transformants were selected by culturing on YPD solid medium containing 100 μg/ml nouseothricin (Jena Bioscience, Germany; indicated as YPDNAT), YPD solid medium with 100 μg/ml hygromycin B (Sigma-Aldrich, USA; indicated as YPDHyB), or YPD solid medium with 100 μg/ml G418 (Duchefa, Netherlands; indicated as YPDNEO). For capsule induction, C. neoformans cells were cultured in liquid Sabouraud dextrose medium (Difco) at 30 °C for 16 h and incubated in 10% Sabouraud dextrose medium (pH 7.3) supplemented with 50 mM morpholinepropanesulfonic acid (MOPS) at 30 °C for 2 days. For melanin production analysis, the cells were spotted onto an L-DOPA agar plate (7.6 mM L-asparagine monohydrate, 5.6 mM glucose, 22 mM KH2PO4, 1 mM MgSO4·7H2O, 0.5 mM L-DOPA, 0.3 mM thiamine-HCl, and 20 nM biotin) and incubated at 30 °C or 37 °C for 2 days.

Transmission electron microscopy

C. neoformans cells were cultured either at an initial optical density of 600 nm (OD600) of 0.2 at 30 °C in YPD medium until OD600 reached 0.8, or for 2 days in 10% Sabouraud media at 30 °C. The cell pellets were washed twice in PBS and fixed for 12 h in 2% glutaraldehyde-2% paraformaldehyde in 0.1 M phosphate buffer (pH 7.4). The cells were washed in 0.1 M phosphate buffer and post-fixed with 1% OsO4 in 0.1 M phosphate buffer for 2 h. The cells were dehydrated using an ascending ethanol series (50, 60, 70, 80, 90, 95, 100%) for 10 min each and infiltrated with propylene oxide for 10 min. The specimens were embedded with a Poly/Bed 812 kit (Polysciences Inc., USA) and polymerized in an electron microscope oven (TD-700, DOSAKA, Japan) at 65 °C for 12 h. The block was cut into 200-nm semi-thin sections with a diamond knife in the ultramicrotome and stained with toluidine blue for observation using optical microscopy. The region of interest was further cut into 80 nm thin sections using the ultramicrotome, placed on copper grids, double stained with 3% uranyl acetate for 30 min and 3% lead citrate for 7 min, and imaged using a transmission electron microscope (JEM-1011, JEOL, Tokyo, Japan) equipped with a Megaview III CCD camera (Soft imaging system-Germany) at the acceleration voltage of 80 kV.

Capsule transfer and shedding analysis

Capsule transfer assays were performed as described previously (Reese and Doering, 2003). Briefly, conditioned medium (CM) was prepared as a source of GXM by culturing the respective strains for 5 days in YPD medium, followed by filtering and storing the culture supernatant at 4 °C. The acceptor strains were cultured overnight at 30 °C in YPD medium. In total, 2 × 106 acceptor cells were incubated with 1 µl of CM for 1 h at room temperature under rotation (18 rpm) and washed twice with PBS. Capsule acquisition was visualized by incubating the cells with an anti-GXM (18B7) antibody conjugated with AlexaFluor 488 (ThermoFisher Scientific, USA) for 1 h at 37 °C, and observing under an Eclipse Ti-E fluorescence microscope (Nikon, Japan), equipped with a Nikon DS-Qi2 camera. The images were processed using the NIS-elements microscope imaging software (Nikon, Japan). Capsule shedding analysis was performed using a modified previously described protocol (Yoneda and Doering, 2008). The respective strains were cultured in 10% Sabouraud media for 2 days, and the culture supernatant was sterile filtered. Enzyme denaturation was performed by subjecting the filtrate to heating at 70 °C for 15 min, followed by centrifugation at 13,000 rpm for 3 min. Intracellular GXM analysis was performed by resuspending the pellets in TNE buffer (50 mM Tris-HCl [pH 7.5], 150 mM NaCl, 5 mM EDTA) with the same volume of glass beads (425–600 µm in diameter, Sigma-Aldrich, USA). The cells were disrupted four times for 15 s at 5,000 rpm using a Precellys 24 Tissue Homogenizer (Bertin Technologies, France), followed by centrifugation at 16,000 rpm for 5 min at 4 °C. Next, 10 µl of the supernatant of either secreted or intracellular polysaccharide was mixed with 6X loading dye and run on a 0.6% certified megabase agarose (Bio-rad, USA) gel in 0.5X TBE (44.5 mM Trisma base, 44.5 mM boric acid, 1 mM EDTA [pH 8.0]) at 25 V for 16 h. The polysaccharides were transferred onto a nylon membrane using the southern blotting technique. The membrane was air-dried, blocked using 5% skim milk, and treated overnight with 2 µg/ml 18B7 antibody. After washing, the membrane was incubated with an anti-mouse peroxidase-conjugated secondary antibody and subjected to detection using chemiluminescence.

Detection of enzymatic activities in intracellular and secretary fractions

Biochemical enzymatic activities of acid phosphatase, urease, and laccase in the intracellular and secreted fractions were assayed spectrophotometrically, using previously described methods (Lev et al., 2014; Fu et al., 2018; de Sousa et al., 2022). Acid phosphatase activity was assayed by culturing cells in MM-KCL medium (0.5% KCl, 15 mM glucose, 10 mM MgSO4·7H2O, 13 mM glycine, and 3 µM thiamine) for 3 h at 30 °C. The culture supernatant and soluble cell lysate were allowed to react with 2.5 mM p-nitrophenyl phosphate (pNPP) for 30 min at 37 °C. Urease activity was determined after culturing the cells in Rapid Urea Broth (RUH broth) for 24 h at 30 °C and subjecting the culture supernatant and soluble cell lysate to incubation with phenol red. Laccase activity was assayed by culturing the cells in asparagine salts media (7.6 mM L-asparagine, 0.1% glucose, 22 mM KH2PO4, 1 mM MgSO4.7H2O, 0.3 mM thiamine-HCl, and 20 nM biotin) for 48 h at 30 °C and allowing the supernatant and soluble cell lysate to react overnight with 10 mM L-DOPA. The reactions were quantified by measuring OD420 (acid phosphatase), OD570 (urease), or OD480 (laccase). Cellulase activity was determined by culturing the cells in cellulose media (0.1% NaNO3, 0.1% K2HPO4, 0.1% KCl, 0.5% MgSO4, 0.5% yeast extract, 0.1% glucose, and 0.5% low viscosity carboxymethyl cellulose) for 24 h at 30 °C. The culture supernatants were concentrated using an Amicon tube (30 kDa cutoff, Sigma-Aldrich, USA). The concentrated supernatants and soluble cell lysates were allowed to react at 40 °C for 10 min with the substrate 4,6-O-(3-ketobutylidene)-4-nitrophenyl-β-D-cellopentaoside (BPNPG5), which was provided in the cellulase assay kit (CellG5 Method, Megazyme, Ireland). The reaction was terminated by adding 2% [w/v] Tris buffer (pH 10), and the absorbance of 4-nitrophenol was measured at 400 nm. Cellulase activity (CellG5 Units/ml) was calculated as indicated in the cellulase assay kit. α-amylase activity was measured using the α-Amylase Activity Colorimetric Assay Kit (Biovision Technologies, USA) according to the manufacturer’s instructions. Yeast cells were cultured overnight in YPD medium at 30 °C, and the culture supernatants were concentrated using an Amicon tube (30 kDa cutoff, Sigma-Aldrich, USA). The concentrated supernatants and soluble cell lysates were incubated for 1 h at 25 °C in assay buffer with the substrate ethylidene-pNP-G7, which was provided in the kit. OD405 was measured. All the activity analysis results were normalized according to cell density (OD600).

Purification, nanoparticle tracking analysis (NTA), and Cryo-TEM imaging of extracellular vesicles

EV purification was performed according to a previously published protocol (Reis et al., 2019) with modifications. One loop of cells was inoculated into 10 ml of liquid YPD and incubated at 30 °C for 24 h with shaking (220 rpm). The cells were washed twice with 10 ml of PBS, counted, and diluted in PBS to a density of 3.5 × 107 cells/ml. Aliquots of cell suspension (300 μl) were spread onto synthetic dextrose (SD) solid medium plates and incubated for 24 h at 30 °C. The cells were carefully recovered from each plate using an inoculation loop, gently resuspended in 30 ml PBS, and pelleted through centrifugation at 4,000 rpm for 10 min at 4 °C. The supernatant was collected and centrifuged again at 15,000 ×g for 15 min at 4 °C. Then, the supernatant was filtered through 0.45-μm syringe filters and ultracentrifuged at 100,000 ×g for 1 h at 4 °C (MLA-50 fixed angle rotor, Beckman Coulter, Germany). The supernatant was discarded, and the EV pellets were collected and resuspended in 1 ml of PBS for immediate use or stored at −80 °C for further experiments. The samples were diluted 100-fold, and EV sizes were measured using an NTA instrument (NFEC-2024-03-295455, Nanosight Pro, Malvern Panalytical, Netherlands) coupled to a 532-nm laser (Malvern Panalytical, Netherlands), SCMOS camera (Hamamatsu Photonics. Japan), and syringe pump (Malvern Panalytical, Netherlands). The data were analyzed using the NS Xplorer software (v1.1.0.6, Malvern Panalytical, Netherlands).

Cryo-TEM imaging was performed by loading purified EVs (3 µl) onto a lacey carbon grid (Lacey Carbon, 300mesh Cu, Ted Pella Inc., USA), which was glow discharged at 15 mA for 60 s. The sample-loaded grid was blotted for 3 s at 15 °C and 100% humidity and immediately plunge-frozen in liquid ethane. The process was performed by Vitrobot Maek IV (Thermofisher Scientific, USA, SNU, CMCI). The frozen grids were imaged using TEM (JEM-2100F, JEOL, Japan); the temperature of the grid was maintained at approximately −180 °C at an acceleration voltage of 200 keV. The images were recorded using an ultrascan 1000 electron detector.

Proteomic analysis based on mass spectrometry and data processing

The samples for proteomic analysis were prepared as described in Supplementary Information. The labeled peptides (total, 100 µg proteins) were combined prior to offline basic reverse-phase liquid chromatographic (bRPLC) fractionation. Linear gradient was performed using buffer A (10 mM TEAB in water) and buffer B (10 mM TEAB in 90% acetonitrile), and 10 fractions were analyzed in total using am LC-MS/MS system. The samples were dissolved in 0.1% formic acid using an UltiMate 3000 RSLCnano system and analyzed using an Orbitrap Eclipse Tribrid mass spectrometer (Thermo Fisher Scientific, USA). All MS raw files were converted into mzML and ms2 file formats using the MSConvert (version 3.0.20033) software. The C. neoformans proteome was determined by downloading a protein FASTA file from Uniprot (http://uniprot.org), which included 7,492 reviewed (Swiss-Prot) and unreviewed (TrEMBL) proteins entries. A proteome search database with reversed sequence and contaminants in the Integrated Proteomics Pipeline (IP2, version 5.1.2, Integrated Proteomics Applications Inc., San Diego, CA) was generated. The proteome search from 20 ms2 files was performed with IP2 and its following parameters, followed by evaluation for a false discovery rate (FDR) using IP2 and Proteininferencer (version 1.0, Integrated Proteomics Applications Inc., USA). Protein quantification and statistical analysis for discovery of differentially expressed proteins (DEPs) was performed using the ms2 files with tandem mass tag (TMT) reporter ions using an in-house program coded using Python 3.8, where t-test and Pearson’s correlation analysis between comparison samples was performed using the scikit-learn (version 0.23.2), Scipy (version 1.6.0), and statsmodels (0.12.1) Python libraries.